\chapter{Conclusions, Limitations, and Future}
Conclusions, Limitations, and Future

Work

Chapter 6

Conclusions,

Limitations, and

Future Work


\section{Introduction to the Chapter}
This chapter closes the GreenLab project report by evaluating the system against the

eight objectives defined in Section 1.4, consolidating the limitations documented

throughout Chapters 3 through 5 into a single account, and proposing a concrete

direction for continued development. Chapter 5 traced individual functional and non-

functional requirements to their implementation evidence (Section 5.6) and recorded

seven specific implementation deviations as they were discovered (Section 6.4). This

chapter operates at the level of the project as a whole: it evaluates whether GreenLab

achieved what it set out to achieve, states its limitations plainly, and proposes the

roadmap that would carry it from a demonstrated prototype to a validated deployment.


\section{Achievement of Project Objectives}
Table 6.1 evaluates each objective from Section 1.4 against the evidence accumulated

across Chapters 3--5, principally the requirements traceability matrix in Section 5.6.

\begin{itemize}
	\item \textbf{Objective 1 --- Layered four-tier architecture with clear interface contracts} (Achieved). The hardware, AI, backend, and dashboard tiers communicate exclusively through the defined REST interfaces described in Section 4.1 and exercised in Chapter 5: the ESP32 firmware never bypasses the backend to reach the AI service or dashboard directly, and each tier's internal implementation (§5.2--5.5) can be replaced without altering another tier's contract.

	\item \textbf{Objective 2 --- Functional and non-functional requirements in formal IEEE 830 style} (Achieved): Sections 3.5 and 3.6 specify 135 functional requirements across nine functional areas and six non-functional quality dimensions. The Section 5.6 traceability matrix confirms every sampled requirement traces to implementation evidence or an explicitly recorded gap.

	\item \textbf{Objective 3 --- Multimodal occupancy detection combining embedded sensors and YOLO-based vision} (Achieved): The Section 5.7 worked trace demonstrates HC-SR04, DHT11, and LDR readings combining with the YOLO service's tracked person count within a single automation decision (FR-OD-01, FR-OD-02: Implemented).

	\item \textbf{Objective 4 --- Lighting control via a decision matrix} (Achieved): The LDR-zone x occupancy-classification matrix (Table 4.6) is implemented as a data-driven rule lookup, satisfying the auditability and configurability goal stated in Objective 4.

	\item \textbf{Objective 5 --- Ventilation control via a fan activation table} (Achieved): The temperature-zone x occupancy-classification table (Table 4.7) follows the same pattern as the lighting matrix and is exercised directly in the Section 5.7 worked example.

	\item \textbf{Objective 6 --- Secure ASP.NET Core REST API with authentication, RBAC, and audit logging} (Achieved): FR-AC-01 through FR-AC-04 are marked Implemented in the Section 5.6 traceability matrix, supported by JWT-based authentication (§4.7.1), a three-role access model (§4.7.2), and an append-only audit log (§4.7.3), each demonstrated against concrete request/response evidence in Chapter 5.

	\item \textbf{Objective 7 --- Angular dashboard for monitoring, override, scheduling, and administration} (Achieved): Sections 5.4--5.5 document eight distinct dashboard views spanning all four required capabilities, with FR-MD-01 and FR-MD-03 both marked Implemented. One presentational inconsistency --- a third device-state pill rendered for a currently-unwired relay channel --- is recorded as a limitation in Section 6.4 and does not affect the four core capabilities this objective requires.

	\item \textbf{Objective 8 --- Complete, print-ready project report by this revision} (Achieved): The report's figures, tables, and formatting satisfy Objective 8. The report was previously incomplete against its own six-chapter structure (no Chapter 6) and carried an unresolved citation-numbering placeholder in Chapter 2; both are resolved as part of this revision.
\end{itemize}

Seven of the eight objectives were fully achieved by the delivered system with no

qualification beyond the implementation-level deviations recorded in Section 6.4.

Objective 8, concerning the report itself, is brought to completion by this chapter and the

Chapter 2 correction above.


\section{Summary of Project Contribution}
GreenLab set out to close a specific gap identified in the literature review (Section 2.11):

no reviewed source combined occupancy-aware automation with role-based governance,

authenticated override, and audit accountability in a single system designed for the

operational realities of a shared academic laboratory. The Table 2.1 comparison matrix

positions GreenLab as the only reviewed system offering multimodal sensing, decision-

matrix-driven automation, and a governed, auditable control layer simultaneously. The

as-built system demonstrates that this combination is achievable within a four-tier

architecture where each tier owns a narrow, well-defined responsibility, connected

through authenticated REST interfaces rather than direct point-to-point coupling. The

Section 5.7 worked example demonstrates the complete path from a physical entry event

to an audited, authorised lighting decision, executing in under 600 milliseconds end to

end, supporting the practical feasibility of the design.


\section{Limitations}

\subsection{Implementation Deviations}
Seven deviations between the Chapter 3/4 specification and the as-built system were

discovered during implementation. They are documented here as formal project limitations.

\begin{itemize}
	\item \textbf{Firmware thresholds hardcoded} (NFR-MAIN-01): LIGHT\_THRESHOLD (2000) and the fan-on temperature (28\textdegree{}C) are compile-time constants. Changing a threshold requires reflashing the device. No /api/devices/\{id\}/config endpoint exists yet. \textit{Recommended fix:} Add a config endpoint; call it at boot; cache in EEPROM for offline use.

	\item \textbf{Two of four relay channels unwired} (Architecture \S4.2): relay3On and relay4On are parsed and logged but not connected to GPIO outputs. Only RELAY\_PIN and RELAY\_PIN2 are physically wired. Hardware not yet expanded beyond two channels. \textit{Recommended fix:} Add \#define RELAY\_PIN3/4; add actuation logic --- no backend change needed.

	\item \textbf{Local fallback re-evaluates rather than freezes} (FR-FH-02): Fallback logic re-evaluates the LDR reading rather than holding the last backend-commanded state as FR-FH-02 specifies. Designed for simplicity over strict spec compliance. \textit{Recommended fix:} Store last backend-commanded light state; restore it on fallback rather than re-evaluating.

	\item \textbf{Teaching Assistant role unmodelled} (FR-AC-02, FR-AC-03): The role was added during implementation but is absent from the Chapter 3 use case model. Discovered during implementation as a real organisational need. \textit{Recommended fix:} Add Teaching Assistant as a named actor in the use case diagram and stakeholder analysis.

	\item \textbf{Occupancy outside session alert unspecified} (Fault handling): The dashboard raises this alert when occupancy is detected outside scheduled hours. No FR ID exists for it. Emergent feature identified during dashboard implementation. \textit{Recommended fix:} Write back as FR-FH-04 in a revised Chapter 3.

	\item \textbf{Device identity compiled into firmware} (NFR-MAIN-01): labId and deviceId are integer literals in the firmware. Deploying to a second lab requires recompiling. No provisioning API was designed for device self-registration. \textit{Recommended fix:} Read labId/deviceId from EEPROM, provisioned via QR code or BLE commissioning.

	\item \textbf{FRAME\_SKIP is count-based, not time-based} (NFR-PERF-03): At variable camera frame rates the effective inference cadence varies. At 15 fps, FRAME\_SKIP=3 gives 5 inferences/s; at 5 fps it gives only 1.7/s. \textit{Recommended fix:} Replace FRAME\_SKIP with a minimum-elapsed-time gate based on time.monotonic().
\end{itemize}


\subsection{Project-Level Limitations}
Beyond the seven implementation-level deviations above, the project carries limitations

that apply to its overall scope and maturity rather than to any single component:

* No field deployment or empirical validation. As stated in the project scope

(Section 1.6), GreenLab has not been installed in a live laboratory environment,

and no energy savings, detection accuracy, or latency figures are reported as

measured GreenLab results beyond the single traced example in Section 5.7.

* No multi-user or extended-duration testing. The system has been exercised

through individual workflow traces and integration checkpoints (Section 1.5), but

not through sustained multi-day operation or concurrent multi-instructor usage.

* Single-prototype hardware scale. All implementation evidence in Chapter 5

reflects one ESP32 unit and one camera source; behaviour under multiple

concurrent laboratory deployments against a shared backend has not been

exercised.


\section{Future Work}
The following four-phase roadmap is proposed as the natural continuation of the project.

Phase 1 --- Configuration and provisioning (near-term). Replace compile-time

firmware thresholds and compiled-in device identity with a runtime configuration and

provisioning workflow: a /api/devices/\{id\}/config endpoint fetched at boot, and a QR-code

or BLE commissioning step for labId/deviceId assignment. This resolves NFR-MAIN-01

without requiring firmware rebuilds for calibration or deployment changes.

Phase 2 --- Requirements and behavioural closure (near-term). Formally add the

Teaching Assistant role to the Chapter 3 use case model and stakeholder analysis, specify

the "occupancy outside session" alert as FR-FH-04, replace the frame-count-based

FRAME\_SKIP sampling gate with a time-based one (resolving NFR-PERF-03), and correct

the network-fallback logic to hold rather than re-evaluate the last commanded relay state

(resolving FR-FH-02). These four items close every specification-level gap recorded in

Section 6.4 without requiring new architecture.

Phase 3 --- Hardware completion (mid-term). Wire the two currently-unused relay

channels to additional laboratory circuits, bringing the physical hardware in line with

the four-channel contract already modelled in the backend response schema, and

expand the hardware node to a small pilot batch (5--10 units) to validate the provisioning

workflow from Phase 1 under realistic multi-device conditions.

Phase 4 --- Field pilot and empirical evaluation (long-term). Deploy GreenLab in a

defined set of pilot laboratories and collect the measurements explicitly excluded from

this report's scope: detection accuracy under real lighting and occupancy conditions,

measured energy consumption before and after deployment, and instructor-reported

usability feedback. This phase converts the project's design-time claims into empirically

supported results and is the single highest-value step toward a production-ready system.

Beyond the immediate four-phase roadmap, several longer-term research directions

warrant consideration. Lightweight model architectures for visual occupancy detection

continue to evolve, with multi-stream deep neural network approaches \cite{mutiset2022}

and optimized convolutional feature map scaling layers \cite{park2024} oﬀering promising

directions for improving inference speed on edge hardware. Self-supervised feature extraction

for panoramic ﬁsheye rectiﬁcation \cite{rodriguez2024} could enable deployment across

a wider range of camera geometries without manual calibration. On the hardware side,

dynamic smart schedule adaptors \cite{kim2024} and decentralized master-slave

architectures \cite{nielsen2025} could extend GreenLab's automation engine to handle

more complex institutional scheduling requirements and larger laboratory estates.


