% ============================================
% Chapter: Implementation
% ============================================

\chapter{Implementation}
\label{ch:implementation}



Implementation

\section{Introduction}


This chapter presents the fully as-built GreenLab system: every line of source code that
runs on each tier of the architecture deﬁned in Chapter 3, paired with annotated screenshots
of the running administration and monitoring interface. Where Chapter 3 ﬁxed what the
system must do and Chapter 4 ﬁxed how it would be structured, Chapter 5 is the proof that
the structure was realised in working software. Every code listing that follows is taken verbatim from the project's committed source ﬁles; every screenshot is from the live deployed
dashboard, not a mock-up or wireframe.

GreenLab is a four-tier IoT system for smart energy management in university computerscience laboratories. The four tiers are: a hardware layer running on the ESP32 microcontroller, a Python AI processing service performing real-time person detection and tracking,
a .NET Core REST API that hosts the business logic and database access layer, and an Angular web dashboard through which administrators and instructors monitor and conﬁgure
the system. This chapter covers the ﬁrst two tiers in full source-code depth --- ﬁrmware
(Section 5.2) and the Python AI service (Section 5.3) --- and covers the dashboard interface
(Sections 5.4 and 5.5) through annotated screenshots tied to functional requirements.

The chapter is structured as follows:


\begin{itemize}
  \item Section 5.2 covers the ESP32 ﬁrmware in six subsections, each presenting a com-
plete function or logical block with line-by-line commentary, design rationale, and
requirement tracebacks.
\end{itemize}

\begin{itemize}
  \item Section 5.3 covers the Python AI processing service in two stages: ﬁrst the single-
camera prototype (cout\_code.txt), then the production multi-camera tracking API
(droidip.py and camera\_sources.py).
\end{itemize}



\begin{itemize}
  \item Section 5.4 covers every screen of the Angular administration interface with anno-
tated screenshots.
\end{itemize}


\begin{itemize}
  \item Section 5.5 covers the monitoring dashboard.
\end{itemize}


\begin{itemize}
  \item Section 5.6 provides a complete requirement traceability matrix.
\end{itemize}


\begin{itemize}
  \item Section 5.7 walks one complete end-to-end motion event through all four tiers with
real JSON payloads.
\end{itemize}


\begin{itemize}
  \item Section 5.8 documents all known implementation gaps and technical debt.
\end{itemize}


\begin{itemize}
  \item Section 5.9 summarises the chapter.
\end{itemize}



Two editorial decisions are stated up front. First, Wi-Fi credentials and device passwords that appear as plain string literals in the ﬁrmware are redacted in every listing in
this chapter; only the surrounding logic that consumes them is shown. Second, this chapter is honest about the small number of places where the as-built code diverges from the
Chapter 3 design --- for example, the ﬁrmware's API paths use /api/devices/ rather than
/sensor/readings, and relay channels three and four are parsed but not yet physically
wired. These deviations are reported explicitly, not smoothed over.



\section{Hardware Layer Implementation --- ESP32 Firmware}


The hardware layer is a single Arduino sketch, arduino\_code.txt, that runs on an ESP32
microcontroller permanently installed in each smart laboratory. The sketch is 487 lines long
and implements six coherent responsibilities: including the required libraries and declaring
all hardware constants (Section 5.2.1); reading the three sensors (Section 5.2.2); responding to direct HTTP commands from the dashboard (Section 5.2.3); authenticating with the
backend using JSON Web Tokens (Section 5.2.4); packaging sensor data and communicating it to the backend on motion events (Section 5.2.5); and running the main event loop that
ties all of the above together (Section 5.2.6). Each section below presents the relevant code
in full, followed by a detailed explanation.




\subsection{Library Includes, Hardware Pin Deﬁnitions, and Global Constants}


The top of the sketch begins with ﬁve library includes and a WebServer instance, followed
by all \#define macros for sensor and relay pins, all tunable constants, and all global vari-
ables used across functions.
1 \#include <WiFi.h>
2 \#include <HTTPClient.h>
3 \#include <DHT.h>
4 \#include <ArduinoJson.h>
5 \#include <WebServer.h>

7 WebServer server(80);



\textit{Listing 5.1: Library includes and local web server declaration (arduino\_code.txt, lines 4--10).}



Each library serves a speciﬁc architectural role. WiFi.h provides the ESP32's Wi-Fi
stack, exposing the WiFi.begin() and WiFi.status() primitives used in setup() and
in the loop's connectivity guard. HTTPClient.h provides the high-level HTTP client used
for both the token-exchange POST and the motion-event POST; it handles TCP connection
management, keep-alive, and response body buffering internally, which would otherwise
require hundreds of lines of socket code. DHT.h is the Adafruit DHT sensor library that
abstracts the one-wire protocol used by the DHT11; without it, the ﬁrmware would need to
implement precise microsecond-level bit-banging for temperature and humidity acquisition.
ArduinoJson.h is Benoît Blanchon's widely-used embedded JSON library; it provides a
document-oriented parser and serialiser that avoids dynamic heap allocation by working
\begin{lstlisting}
from a statically-sized DynamicJsonDocument or StaticJsonDocument. WebServer.h
\end{lstlisting}

wraps the ESP32's built-in TCP server and provides the server.on() route-registration
and server.handleClient() polling API used for the local light and fan control end-
points described in Section 5.2.3.


Design note. WebServer runs on port 80 in addition to the main backend communication.
This gives the Angular dashboard a direct HTTP path to toggle the light and fan without
sending a round-trip through the .NET Core backend --- useful for manual override during
demos or when the backend is unrestarting.




1 \#define TRIG\_PIN 4 // HC-SR04 ultrasonic trigger output
2 \#define ECHO\_PIN 2 // HC-SR04 ultrasonic echo input
3 \#define RELAY\_PIN 26 // Relay channel 1 -- controls laboratory lights
4 \#define RELAY\_PIN2 27 // Relay channel 2 -- controls laboratory fan
5 \#define LDR\_PIN 34 // Analogue input for LDR (light-dependent resistor)
6 \#define DHT\_PIN 5 // Digital data pin for DHT11 temperature/humidity sensor
7 \#define DHT\_TYPE DHT11


\textit{Listing 5.2: GPIO pin assignments for all sensors and relay channels (arduino\_code.txt, lines 11--}

17).


Pin selection is not arbitrary. GPIO 34 is chosen for the LDR because it is one of the
ESP32's input-only ADC pins --- it cannot be accidentally conﬁgured as an output. GPIO
4 for TRIG\_PIN and GPIO 2 for ECHO\_PIN are general-purpose I/O pins; GPIO 2 is also
connected to the ESP32's built-in LED on most development boards, which gives a visual
ﬂicker on every ultrasonic pulse during debugging. GPIO 26 and 27 are DAC-capable pins
on the ESP32, but they are used here purely as digital outputs to drive the relay module's
control inputs; no analogue output is needed. GPIO 5 for DHT\_PIN is chosen because it is
one of the few pins that does not have a boot-mode conﬂict on the ESP32.
1 int labId = 1;
2 int deviceId = 1;


4 const char* Devicepassword = "[REDACTED]";
5 const char* ssid = "[REDACTED]";
6 const char* password = "[REDACTED]";

8 const char* serverUrl = "http://192.168.1.7:5076/api/devices/1/event";
9 const char* lightApi = "http://192.168.1.7:5076/api/devices/light";
10 const char* TokenGen = "http://192.168.1.7:5076/api/devices/token";

12 String jwtToken = "";
13 unsigned long tokenExpireMillis = 0;

15 static bool lastLightState = false;
16 static bool lastFanState = false;


18 const int DISTANCE\_THRESHOLD = 50; // centimetres
19 const unsigned long COOLDOWN = 2000; // milliseconds




20 const int LIGHT\_THRESHOLD = 2000; // LDR ADC value
21 const unsigned long LIGHT\_INTERVAL = 5000; // ms (reserved for future periodic
checks)

23 bool wasNear = false;
24 unsigned long lastToggleTime = 0;


26 DHT dht(DHT\_PIN, DHT\_TYPE);


\textit{Listing 5.3: Device identity, network endpoints, JWT state, relay state trackers, and tuning constants}

(arduino\_code.txt, lines 19--46).


labId and deviceId are the identiﬁers that the backend uses to look up which lab-
oratory this ESP32 belongs to and which device record to authenticate against. They are
compiled-in integer literals in the current build, which means reﬂashing is required to de-
ploy the same ﬁrmware binary to a second laboratory with a different ID --- a maintainabil-
ity gap noted in Section 5.8 and tracked against NFR-MAIN-01.

The three API endpoint strings embed the local network IP address 192.168.1.7 and
port 5076, which are the host:port of the .NET Core backend process on the development
network. serverUrl points to the motion-event endpoint; lightApi points to an endpoint
used for direct light-state queries (reserved for future use); TokenGen points to the device
authentication endpoint.

jwtToken and tokenExpireMillis implement the client side of the JWT caching
strategy: after a successful token exchange, jwtToken holds the bearer token string and
tokenExpireMillis holds the millis() value at which the ﬁrmware will proactively
refresh it. lastLightState and lastFanState are the differential-update guards that
prevent unnecessary relay switching chatter --- a relay is only toggled when the commanded
state differs from the last applied state.

DISTANCE\_THRESHOLD (50 cm) deﬁnes the boundary within which the HC-SR04 con-
siders a person to be ``near'' the sensor. COOLDOWN (2000 ms) is the minimum gap between
consecutive motion-event POSTs, preventing the ﬁrmware from ﬂooding the backend if a
person lingers near the sensor. LIGHT\_THRESHOLD (2000) is the ADC reading below which
the local fallback logic considers the room too dark to leave the light off. LIGHT\_INTERVAL
is declared but currently unused --- it was intended for a periodic light-level check indepen-



dent of motion events, which remains as planned future work.

Table 5.1: Firmware constants and their alignment with the Add Lab form defaults.

Constant Value Purpose Conﬁgurable via Add
Lab?
50 cm
DISTANCE\_THRESHOLD Deﬁnes ``presence'' zone for No --- hardcoded (see
HC-SR04 §5.8)
COOLDOWN 2000 ms Minimum gap between mo- No --- hardcoded
tion reports
LIGHT\_THRESHOLD 2000 LDR threshold for local No --- must match Add
(ADC) fallback Lab form defaults man-
ually
LIGHT\_INTERVAL 5000 ms Reserved for periodic light N/A --- unused in cur-
checks rent build


The table above conﬁrms the alignment between ﬁrmware constants and the Add Lab
form defaults (Section 5.4.1): the values were set consistently by hand on both sides during
integration, not by the ﬁrmware reading them from the backend API.


\subsection{Sensor Reading Functions}


Three sensor-reading functions are deﬁned before any of the communication or authenti-
cation logic. Keeping them short and side-effect-free makes the main loop's intent clear:
sensors are read, a decision is made, and then --- only if necessary --- a network request is
sent.
1 // HC-SR04 Ultrasonic Distance Measurement
2 long readDistanceCM() {
3 // Step 1: ensure trigger is low for a clean pulse
4 digitalWrite(TRIG\_PIN, LOW);
5 delayMicroseconds(2);

7 // Step 2: fire a 10-microsecond HIGH pulse on the trigger pin
8 digitalWrite(TRIG\_PIN, HIGH);
9 delayMicroseconds(10);
10 digitalWrite(TRIG\_PIN, LOW);

12 // Step 3: measure the duration of the echo pulse
13 // timeout of 30000 us corresponds to ~5 m -- beyond lab sensor range
14 long duration = pulseIn(ECHO\_PIN, HIGH, 30000);





16 // Step 4: guard against timeout (no echo returned)
17 if (duration == 0) return -1;

19 // Step 5: convert travel time to distance
20 // sound travels 0.034 cm/us; divide by 2 for one-way trip
21 return duration * 0.034 / 2;
22 }


\textit{Listing 5.4: readDistanceCM(): HC-SR04 ultrasonic distance measurement with timeout guard}

(arduino\_code.txt, lines 48--59).


The HC-SR04 works by emitting eight 40 kHz ultrasonic pulses when the trigger pin is
held HIGH for at least 10 microseconds. The sensor then raises the echo pin HIGH for a du-
ration proportional to the round-trip travel time of the ﬁrst reﬂected pulse. pulseIn(ECHO\_PIN,
HIGH, 30000) blocks until the echo pin goes HIGH and then measures how long it stays
HIGH, with a 30,000-microsecond timeout. The 30 ms timeout is chosen to correspond to a
maximum measurable distance of approximately 510 cm --- well beyond the 50 cm thresh-
old the ﬁrmware cares about, ensuring the timeout only ﬁres when there is genuinely no
reﬂecting object in range. The -1 sentinel return on timeout allows the caller to distinguish
``no target'' from a legitimately short distance of zero centimetres.

The conversion formula --- duration * 0.034 / 2 --- derives from the speed of
sound at room temperature being approximately 340 m/s = 0.034 cm/µs. Dividing by two
accounts for the signal travelling to the target and back.


Physics note. At 26$^\circ$C (the temperature logged in the Section 5.7 worked example), the
speed of sound is approximately 346 m/s rather than the assumed 340 m/s, introducing a
$\sim$1.8% measurement error. For a 50 cm threshold this is a 0.9 cm error --- negligible for
presence detection but worth noting for any future use case that requires precise ranging.
1 // LDR Ambient Light Reading
2 int readLight() {
3 return analogRead(LDR\_PIN);
4 }


\textit{Listing 5.5: readLight(): single-line LDR analogue read returning 0--4095 on the ESP32's 12-bit}

ADC (arduino\_code.txt, line 62).



readLight() is intentionally a single-expression function. The ESP32's ADC is 12-
bit, returning values from 0 (no voltage) to 4095 (3.3 V). A lower LDR reading means more
light is present --- the LDR's resistance falls as light intensity increases, which reduces the
voltage at the analogue pin when the LDR is wired in a voltage-divider conﬁguration with a
ﬁxed resistor. LIGHT\_THRESHOLD = 2000 therefore corresponds to a mid-scale brightness
level; readings below 2000 indicate a well-lit room and readings above 2000 indicate a
darker room. The ﬁrmware's local fallback turns the light on when the reading is above this
threshold (i.e., the room is dark), matching the expected behaviour.

A known limitation of the ESP32's ADC is signiﬁcant non-linearity near 0 V and
\section{V, and sensitivity to crosstalk from Wi-Fi radio activity. The ADC readings are therefore}

treated as a relative brightness indicator rather than a calibrated lux measurement, which
is sufﬁcient for the binary ``room is dark / room is bright'' decision the ﬁrmware needs to
make.

1 // DHT11 temperature and humidity (called in loop())
2 float temp = dht.readTemperature();
3 float humidity = dht.readHumidity();

5 if (isnan(temp) || isnan(humidity)) {
6 Serial.println("Sensor error");
7 return;
8 }



\textit{Listing 5.6: DHT11 read and NaN guard in loop() (arduino\_code.txt, lines 321--328).}



The DHT11 sensor is read inline in loop() rather than in a dedicated function because
both readTemperature() and readHumidity() must be called in the same loop iteration
to remain synchronised. The DHT library internally caches the last successful reading and
can return NaN (Not a Number) when the one-wire protocol communication fails --- a
common occurrence if the sensor is read more frequently than once per two seconds, or
if there is electrical noise on the data line. The isnan() guard causes loop() to return
immediately without sending any network request, avoiding a sensor-error value of NaN
being serialised into the JSON body and causing a deserialization failure on the backend.

The DHT11's rated accuracy is ±2$^\circ$C for temperature and ±5% for relative humidity,
with a measurement range of 0--50$^\circ$C and 20--90% RH. For the fan-control decision (turn



fan on if temperature > 28$^\circ$C), this accuracy is adequate --- the 2$^\circ$C uncertainty means the
fan may switch on slightly early or slightly late, but it will not fail to respond to a genuinely
hot room.



\subsection{Local HTTP Server --- Direct Relay Control Endpoints}


In addition to communicating with the .NET Core backend, the ESP32 runs a lightweight
HTTP server on port 80. This server accepts four routes that allow the Angular dashboard
--- or any HTTP client on the local network --- to directly command the light and fan relays
without going through the backend. This is the manual-override mechanism used during
demonstrations and when the backend is temporarily unavailable.
1 void handleLightOff() {
2 digitalWrite(RELAY\_PIN, HIGH); // HIGH = relay open = light OFF
3 Serial.println("Light OFF from API Button");
4 server.send(200, "application/json",
5 "{\"status\":\"ok\",\"light\":\"off\"}");
6 lastLightState = false; // keep state tracker in sync
7 }

9 void handleLightOn() {
10 digitalWrite(RELAY\_PIN, LOW); // LOW = relay closed = light ON
11 Serial.println("Light ON from API Button");
12 server.send(200, "application/json",
13 "{\"status\":\"ok\",\"light\":\"on\"}");
14 lastLightState = true;
15 }

17 void handleFanOn() {
18 digitalWrite(RELAY\_PIN2, LOW); // LOW = relay closed = fan ON
19 Serial.println("Fan ON from API");
20 server.send(200, "application/json", "{\"fan\":\"on\"}");
21 lastFanState = true;
22 }

24 void handleFanOff() {
25 digitalWrite(RELAY\_PIN2, HIGH); // HIGH = relay open = fan OFF
26 Serial.println("Fan OFF from API");




27 server.send(200, "application/json", "{\"fan\":\"off\"}");
28 lastFanState = false;
29 }


\textit{Listing 5.7: Four direct relay-control HTTP handlers: /light/on, /light/off, /fan/on, /fan/off}

(arduino\_code.txt, lines 64--102).


The relay logic convention used throughout the ﬁrmware deserves an explicit note:
HIGH on a relay control pin opens the relay (disconnects the load), and LOW closes it
(connects the load). This is the active-low convention used by the opto-isolated relay mod-
ule connected to the ESP32. Writing HIGH on RELAY\_PIN therefore turns the light off;
writing LOW turns it on. This is counter-intuitive when reading the code for the ﬁrst time
but is correct for the hardware conﬁguration in use.

Each handler updates the lastLightState or lastFanState tracker immediately
after the GPIO write. This is critical: if the dashboard sends a direct /light/on com-
mand and then, moments later, the main loop processes a new backend response that con-
tains lightOn: false, the differential update guard (lightOn != lastLightState)
will correctly detect the discrepancy and toggle the relay back off. Without updating the
tracker in the handler, the guard would think the light was already off and do nothing,
leaving the relay in a state the backend did not command.

Table 5.2: Local HTTP server routes for direct relay control.

Route Method GPIO write Response body State tracker
updated
/light/on GET RELAY\_PIN → {"status":"ok","light":"on"}
lastLightState
LOW = true
/light/off GET RELAY\_PIN → {"status":"ok","light":"off"}
lastLightState
HIGH = false
/fan/on GET RELAY\_PIN2 → {"fan":"on"} lastFanState
LOW = true
/fan/off GET RELAY\_PIN2 → {"fan":"off"} lastFanState
HIGH = false


These four routes are registered in setup() and polled on every loop() iteration via
server.handleClient(). Because handleClient() is non-blocking --- it processes at
most one pending request per call --- it does not interfere with the motion-detection and
sensor-reading logic that runs in the same loop iteration. The server responds to requests



synchronously within the handler function, so no asynchronous callback or interrupt is
needed.



\subsection{Device Authentication --- JWT Token Lifecycle}


Every HTTP request the ESP32 makes to the .NET Core backend must carry a valid JSON
Web Token issued speciﬁcally to that device. The ﬁrmware manages this token lifecycle
through three cooperating functions: handleTokenResponse() parses a token-exchange
response and stores the result; refreshToken() performs the actual credential exchange
with the backend; and ensureTokenValid() is a lightweight guard called before every
outgoing request to ensure the stored token is still usable.
1 bool handleTokenResponse(String response) {
2 DynamicJsonDocument doc(512);
3 DeserializationError error = deserializeJson(doc, response);
4 if (error) {
5 Serial.println("Token Parse Failed");
6 return false;
7 }


9 jwtToken = doc["token"].as<String>();
10 String expiresAt = doc["expiresAt"].as<String>();

12 Serial.println("Token: " + jwtToken);
13 Serial.println("ExpiresAt: " + expiresAt);

15 // Conservative assumed expiry: 50 minutes from now
16 // Real expiry is in expiresAt but we avoid ISO-8601 parsing on ESP32
17 tokenExpireMillis = millis() + (50UL * 60UL * 1000UL);

19 return jwtToken != "";
20 }


\textit{Listing 5.8: handleTokenResponse(): JWT extraction and conservative 50-minute assumed expiry}

(arduino\_code.txt, lines 104--124).


handleTokenResponse() is called by refreshToken() after a successful HTTP
response. It allocates a 512-byte JSON document on the stack, parses the response body



into it, and extracts two ﬁelds: token (the JWT string to be used as a bearer credential)
and expiresAt (an ISO-8601 timestamp string representing when the server considers the
token expired).

The expiresAt string is read and printed to the serial monitor for diagnostic purposes
but is not parsed into a comparable timestamp value.

The reason for not parsing expiresAt is pragmatic: the ESP32 runs FreeRTOS and
has access to millis() (milliseconds since boot) but does not have a calibrated real-time
clock. Converting an ISO-8601 string like "2026-06-04T09:00:00Z" into a millis()
value would require knowing the current Unix timestamp, which would need an NTP syn-
chronisation step that is not in the current build. Instead, tokenExpireMillis is set to
millis() + 3,000,000 (50 minutes in milliseconds), using unsigned long arithmetic
with the UL sufﬁx to prevent 32-bit overﬂow on the multiplication. This is deliberately
shorter than the server's conﬁgured token lifetime, so the device refreshes early and never
risks sending a request with a token that expired moments ago.



Security note. Although the device password appears in the ﬁrmware source as a plain
string constant, it is transmitted over the local network in a JSON POST body --- not sent as
a query parameter or in a URL. The /api/devices/token endpoint should be restricted to
the local subnet in a production deployment to prevent credential interception from outside
the laboratory network.

1 bool ensureTokenValid() {
2 // Refresh if token is missing (first boot or parse failure)
3 // or if millis() has passed the assumed expiry timestamp
4 if (jwtToken == "" || millis() > tokenExpireMillis) {
5 Serial.println("Token expired or missing -> Refreshing...");
6 return refreshToken();
7 }
8 return true; // token is present and within assumed lifetime
9 }


\textit{Listing 5.9: ensureTokenValid(): lazy refresh guard --- zero network calls when token is fresh}

(arduino\_code.txt, lines 126--133).


ensureTokenValid() is called at the top of sendMotionRequest() before any HTTP



activity. If the token is present and millis() has not yet passed tokenExpireMillis, the
function returns immediately with no network activity --- no HTTP connection is opened,
no socket is allocated. This is the normal path on the vast majority of loop() iterations.
Only when the token is missing or presumed expired does the function call refreshToken(),
which takes hundreds of milliseconds due to the network round-trip.

The millis() overﬂow edge case deserves mention. millis() returns an unsigned
\begin{lstlisting}
long on the ESP32, which overﬂows back to zero after approximately 49.7 days of continu-
\end{lstlisting}

ous uptime. If tokenExpireMillis was set close to the overﬂow boundary and millis()
subsequently overﬂows, the condition millis() > tokenExpireMillis will be false
when it should be true --- causing the ﬁrmware to use a possibly-expired token. For a
laboratory that is powered off overnight (fewer than 49.7 continuous days), this edge case
is irrelevant. For a permanently-powered deployment, this should be addressed by using
elapsed-time arithmetic rather than an absolute timestamp.

1 bool refreshToken() {
2 if (WiFi.status() != WL\_CONNECTED) return false;


4 HTTPClient http;
5 http.begin(TokenGen);
6 http.addHeader("Content-Type", "application/json");


8 DynamicJsonDocument doc(128);
9 doc["labId"] = labId;
10 doc["deviceId"] = deviceId;
11 doc["password"] = Devicepassword; // credential sent in body, not URL


13 String json;
14 serializeJson(doc, json);

16 Serial.println("Token Request:");
17 Serial.println(json);

19 int code = http.POST(json);
20 if (code <= 0) {
21 Serial.println("HTTP failed");
22 http.end();
23 return false;




24 }

26 String response = http.getString();
27 Serial.println("Token response:");
28 Serial.println(response);

30 http.end();
31 return handleTokenResponse(response);
32 }


\textit{Listing 5.10: refreshToken(): full device-credential exchange with the /api/devices/token endpoint}

(arduino\_code.txt, lines 157--191).


refreshToken() opens a new HTTPClient connection to TokenGen, adds the Content-Type
header, builds a 128-byte JSON document containing labId, deviceId, and the device
password, serialises it to a String, and POSTs it. A return code of 0 or negative from
http.POST() indicates a transport-level failure (TCP connection refused, DNS failure,
timeout) and causes the function to return false without attempting to parse an empty
response. A positive return code --- including 401 Unauthorized or 500 Internal Server
Error --- indicates the server responded; the response body is retrieved and passed to
handleTokenResponse(). If handleTokenResponse() fails to ﬁnd a non-empty token
string, it returns false, and refreshToken() propagates that failure to its caller.

The code <= 0 guard is important: on the ESP32, HTTPClient returns negative val-
ues for library-level errors (e.g., HTTPC\_ERROR\_CONNECTION\_REFUSED = -1, HTTPC\_ERROR\_SEND\_HEADE
= -2). Without this guard, attempting to call getString() on a failed request would return
an empty string and handleTokenResponse() would fail to parse it, achieving the same
result but generating confusing debug output. The guard makes the failure path explicit.



\subsection{Motion-Triggered Sensor Reporting --- sendMotionRequest()}


sendMotionRequest() is the ﬁrmware's primary communication function. It is called
once per detected motion event and is responsible for packaging all ﬁve sensor readings
into a JSON body, attaching the JWT, POSTing to the backend, handling a 401 response
by refreshing the token and retrying, and returning the backend's response body for the
relay-actuation logic in Section 5.2.6.




1 String sendMotionRequest(bool motion, long distance, int ldr, float temp, float
hum) {
2 // Guard 1: ensure a valid JWT is cached before any HTTP activity
3 if (!ensureTokenValid()) {
4 Serial.println("Token invalid");
5 return "";
6 }


8 // Guard 2: confirm Wi-Fi is still connected
9 if (WiFi.status() != WL\_CONNECTED) {
10 Serial.println("WiFi not connected");
11 return "";
12 }

14 HTTPClient http;
15 http.begin(serverUrl);
16 http.setTimeout(10000); // 10-second timeout for this request
17 http.addHeader("Authorization", "Bearer " + String(jwtToken));
18 http.addHeader("Content-Type", "application/json");

20 // Build JSON payload manually (avoids another 512-byte DynamicJsonDocument)
21 String json = "{";
22 json += "\"motionDetected\":" + String(motion ? "true" : "false") + ",";
23 json += "\"distanceCm\":" + String(distance) + ",";
24 json += "\"ldrValue\":" + String(ldr) + ",";
25 json += "\"temperatureCelsius\":" + String(temp, 1) + ",";
26 json += "\"humidity\":" + String(hum, 1);
27 json += "}";

29 Serial.println("Sending JSON:");
30 Serial.println(json);

32 int code = http.POST(json);
33 String response = "";


\textit{Listing 5.11: sendMotionRequest() Part A: guards, header setup, JSON body construction, and}

initial POST (arduino\_code.txt, lines 195--236).


The function begins with two pre-ﬂight guards. The ﬁrst calls ensureTokenValid(),



which may itself call refreshToken() and incur a network round-trip if the token has
expired. If token validation fails for any reason, the function returns an empty string im-
mediately without attempting the motion POST --- a fast-fail that prevents a request that
is guaranteed to be rejected. The second guard checks WiFi.status() again even though
ensureTokenValid() also checks it internally, because a race condition is theoretically
possible: the Wi-Fi connection could drop between the token validation check and the
HTTP POST attempt. In practice, on a stable local network this race is extremely unlikely,
but the second guard makes the failure path explicit for diagnostics.

The JSON body is built using manual string concatenation rather than a DynamicJsonDocument
because the structure is ﬁxed and known at compile time: ﬁve named ﬁelds, no nesting, no
arrays. Manual construction avoids allocating another JSON document on the heap and
produces smaller, faster code for this speciﬁc case. The String(temp, 1) call serialises
the ﬂoat temperature to one decimal place, which is the precision the DHT11 can reliably
provide.
1 // 401 Retry: if the backend rejects the token, refresh once and retry
2 if (code == 401) {
3 Serial.println("401 -> Refresh token");
4 if (refreshToken()) {
5 http.end();
6 http.begin(serverUrl);
7 http.addHeader("Authorization", "Bearer " + String(jwtToken));
8 http.addHeader("Content-Type", "application/json");
9 code = http.POST(json);
10 }
11 }

13 if (code > 0) {
14 response = http.getString();
15 Serial.println("Response: " + response);
16 } else {
17 Serial.println("HTTP Request Failed");
18 }

20 http.end();
21 return response;
22 }







\textit{Listing 5.12: sendMotionRequest() Part B: 401 retry logic, response collection, and cleanup}

(arduino\_code.txt, lines 237--264).


The 401-retry block is the ﬁrmware's implementation of self-healing authentication. If
the backend returns HTTP 401, the ﬁrmware calls refreshToken() to obtain a new JWT,
ends the current HTTPClient session (freeing its socket), opens a new session to the same
URL with the new token in the Authorization header, and retries the POST with the same
JSON body. The JSON body string is intentionally declared before the retry block so that
it is available without reconstruction. This retry happens at most once per motion event ---
if the refreshed token is also rejected, the code falls through to the code > 0 check, which
will be false, and the function returns an empty string. A second retry is not attempted, to
avoid an inﬁnite authentication loop in the event of a backend misconﬁguration.

The http.end() call in the retry block is critical: the ESP32's HTTP client internally
holds a TCP socket open after a request. Without calling end() before calling begin()
again, a second connection to the same host would either reuse the existing socket (which
is in an undeﬁned state after a 401 response) or open a second socket and leak the ﬁrst.
Either outcome would eventually exhaust the ESP32's limited socket pool.


\subsection{Main Firmware Logic --- setup() and loop()}


setup() runs once on boot and performs the following initialisation steps in order: pin
mode conﬁguration, relay initialisation to safe-off state, serial monitor at 115200 baud,
DHT sensor initialisation, Wi-Fi connection, initial token acquisition, HTTP route regis-
tration, and local web server start. loop() is called continuously by the Arduino runtime
and handles Wi-Fi reconnection, local server request polling, sensor reading, motion-edge
detection, and relay actuation.
1 void setup() {
2 // Configure sensor and relay pins
3 pinMode(TRIG\_PIN, OUTPUT);
4 pinMode(ECHO\_PIN, INPUT);
5 pinMode(RELAY\_PIN, OUTPUT);
6 pinMode(RELAY\_PIN2, OUTPUT);






8 // Safe initial state: both relays open (light and fan OFF)
9 // HIGH = relay open on active-low relay module
10 digitalWrite(RELAY\_PIN, HIGH);
11 digitalWrite(RELAY\_PIN2, HIGH);

13 Serial.begin(115200);
14 dht.begin();

16 // Block until Wi-Fi connects (dot-print progress indicator)
17 WiFi.begin(ssid, password);
18 while (WiFi.status() != WL\_CONNECTED) {
19 delay(500);
20 Serial.print(".");
21 }
22 Serial.println("\nConnected");
23 Serial.println(WiFi.localIP());


25 // Acquire initial token; log failure but do not halt
26 if (!refreshToken()) {
27 Serial.println("Failed to get token at startup");
28 }

30 // Register direct relay-control routes
31 server.on("/light/on", handleLightOn);
32 server.on("/light/off", handleLightOff);
33 server.on("/fan/on", handleFanOn);
34 server.on("/fan/off", handleFanOff);
35 server.begin();

37 Serial.println("ESP32 Server Started");
38 }


\textit{Listing 5.13:      setup():   full initialisation sequence from pin conﬁguration to server start}

(arduino\_code.txt, lines 267--310).


Setting both relay pins HIGH immediately after pinMode() is the safe-state initiali-
sation: it ensures that even if setup() takes several seconds (due to the Wi-Fi connection
loop), the light and fan are off during that time and do not switch on transiently. The alter-
native --- not writing HIGH until after Wi-Fi connects --- would leave the relay pins in an



indeterminate state during the connection phase, which on some relay modules could cause
a brief activation glitch.

The blocking while (WiFi.status() != WL\_CONNECTED) loop is acceptable in
setup() because the ESP32 has no useful work to do until it has network connectivity
--- all of its functionality depends on being able to reach the backend. In loop(), by con-
trast, the Wi-Fi reconnection logic is non-blocking (see below) to avoid stalling the local
HTTP server.

Calling refreshToken() at the end of setup() pre-populates jwtToken so that
the ﬁrst motion event after boot does not need to wait for a token exchange before it
can POST sensor data. If the token exchange fails at startup (e.g., the backend is not
yet running), a warning is printed but setup() completes normally --- the next call to
ensureTokenValid() in sendMotionRequest() will retry the exchange.
1 void loop() {
2 // --- Sensor reads (every iteration) ---
3 long distance = readDistanceCM();
4 int lightValue = readLight();
5 float temp = dht.readTemperature();
6 float humidity = dht.readHumidity();

8 // DHT guard: abort iteration on sensor error
9 if (isnan(temp) || isnan(humidity)) {
10 Serial.println("Sensor error");
11 return;
12 }


14 bool isNear = (distance > 0 \&\& distance <= DISTANCE\_THRESHOLD);

16 // --- Wi-Fi watchdog (non-blocking) ---
17 static unsigned long lastWifiTry = 0;
18 if (WiFi.status() != WL\_CONNECTED) {
19 if (millis() - lastWifiTry > 5000) {
20 Serial.println("Reconnecting WiFi...");
21 WiFi.disconnect();
22 WiFi.begin(ssid, password);
23 lastWifiTry = millis();
24 }




25 return; // skip network operations until Wi-Fi is back
26 }


28 // --- Service local HTTP requests ---
29 server.handleClient();

31 // --- Motion edge detection with cooldown ---
32 if (isNear \&\& !wasNear \&\& millis() - lastToggleTime > COOLDOWN) {
33 Serial.println("Motion detected -> Sending to API");
34 String response = sendMotionRequest(isNear, distance, lightValue, temp,
humidity);


\textit{Listing 5.14: loop() Part A: sensor reads, NaN guard, Wi-Fi watchdog, server polling, motion edge}

detection (arduino\_code.txt, lines 312--357).


loop() begins with sensor reads on every iteration, even if no motion event will
be reported. This ensures the sensor values are always fresh when needed. The bool
isNear expression combines two conditions: distance > 0 guards against the -1 sentinel
\begin{lstlisting}
returned by readDistanceCM() on timeout (which would incorrectly satisfy distance <=
\end{lstlisting}

DISTANCE\_THRESHOLD), and distance <= DISTANCE\_THRESHOLD implements the actual
50 cm threshold. A value of -1 is not $\le$ 50, but because distance is declared as long
(signed), -1 would satisfy -1 <= 50 --- hence the explicit > 0 guard.

The Wi-Fi watchdog uses a static local variable lastWifiTry, which is initialised
once on the ﬁrst loop() call and retains its value across iterations. Checking millis()
- lastWifiTry > 5000 before calling WiFi.begin() again prevents the ﬁrmware from
hammering the Wi-Fi stack with connection attempts every 100 ms (the delay at the bottom
of loop()). Calling WiFi.disconnect() before WiFi.begin() clears any stale connec-
tion state, which can prevent the ESP32's internal Wi-Fi stack from getting stuck in an
intermediate state after a drop. The return immediately after the watchdog block ensures
that no network operations (sendMotionRequest, server.handleClient) are attempted
\begin{lstlisting}
while the device is ofﬂine.
\end{lstlisting}

1 // --- Local fallback if backend unreachable ---
2 bool apiFailed = (response == "");
3 if (apiFailed) {
4 Serial.println("Fallback to local logic");
5 static bool lastState = false;




6 bool newState = (lightValue < LIGHT\_THRESHOLD);
7 if (newState != lastState) {
8 digitalWrite(RELAY\_PIN, newState ? LOW : HIGH);
9 lastState = newState;
10 }
11 return;
12 }

14 // --- Parse backend JSON response ---
15 DynamicJsonDocument doc(512);
16 DeserializationError error = deserializeJson(doc, response);
17 if (error) {
18 Serial.println("JSON Parse Failed");
19 return;
20 }

22 bool lightOn = doc["lightOn"];
23 bool fanOn = doc["fanOn"];
24 bool relay3On = doc["relay3On"];
25 bool relay4On = doc["relay4On"];
26 bool motionIgnored = doc["motionIgnored"];
27 int occupancy = doc["occupancyCount"];


\textit{Listing 5.15:      loop() Part B: local fallback logic and backend JSON response parsing}

(arduino\_code.txt, lines 358--396).


The fallback branch is triggered when sendMotionRequest() returns an empty string,
meaning the POST failed at transport level (no Wi-Fi, TCP timeout, or token refresh fail-
ure). In this case, the ﬁrmware applies a single local rule: if lightValue < LIGHT\_THRESHOLD,
the room is currently well-lit, so the light stays off; if lightValue >= LIGHT\_THRESHOLD,
the room is dark, so the light is turned on. A static lastState variable tracks the last
locally-applied light state so the relay is only toggled when the LDR reading crosses the
threshold, not on every loop iteration --- the same differential-update pattern used for the
backend-driven relay logic.

The fallback uses only the light relay (RELAY\_PIN), not the fan relay. This is a de-
liberate conservative choice: the local LDR reading can justify a lighting decision (dark
room → light on), but the ﬁrmware has no local logic for fan control without the backend's



temperature rule evaluation. Leaving the fan in its last known state (which was initialised
to HIGH/off in setup()) is the safe-default behaviour for ventilation.


1 // --- Room-empty override: force all relays off ---
2 if (occupancy == 0) {
3 Serial.println("Room Empty -> FORCE OFF");
4 digitalWrite(RELAY\_PIN, HIGH); // light off
5 digitalWrite(RELAY\_PIN2, HIGH); // fan off
6 lastFanState = false;
7 lastLightState = false;
8 lastToggleTime = millis();
9 return; // skip API-driven relay logic below
10 }


12 // --- Relay actuation from API response ---
13 if (lightOn != lastLightState) {
14 digitalWrite(RELAY\_PIN, lightOn ? LOW : HIGH);
15 lastLightState = lightOn;
16 Serial.println(lightOn ? "Light ON" : "Light OFF");
17 }

19 if (fanOn != lastFanState) {
20 digitalWrite(RELAY\_PIN2, fanOn ? LOW : HIGH);
21 lastFanState = fanOn;
22 Serial.println(fanOn ? "Fan ON" : "Fan OFF");
23 }

25 // Relay 3 and 4: parsed and logged -- not yet wired to GPIO
26 if (relay3On) Serial.println("Relay3 ON");
27 else Serial.println("Relay3 OFF");

29 if (relay4On) Serial.println("Relay4 ON");
30 else Serial.println("Relay4 OFF");

32 // motionIgnored: backend chose not to act on this event
33 if (motionIgnored) {
34 Serial.println("Motion Ignored by Server");
35 lastToggleTime = millis();
36 return;




37 }
38 } // end if (isNear \&\& !wasNear \&\& cooldown)


40 wasNear = isNear;
41 delay(100);
42 } // end loop()



\textit{Listing 5.16: loop() Part C: room-empty override, differential relay actuation, motionIgnored}

handling, and loop bookkeeping (arduino\_code.txt, lines 397--445).



The room-empty override is evaluated before the per-relay differential logic. If occupancyCount
is 0, the backend is asserting that the AI service has conﬁrmed no people are present in the
room. In this case, both relays are forced off unconditionally --- even if lightOn or fanOn
in the same response happen to be true (which should not occur from a correctly behaving
backend, but the ﬁrmware does not assume correct backend behaviour). Both state trackers
are reset to false and lastToggleTime is updated to enforce the cooldown before the
next motion event can be processed. The return prevents the differential relay logic below
\begin{lstlisting}
from being evaluated.

\end{lstlisting}

The differential relay actuation pattern --- if (lightOn != lastLightState) ---
is the ﬁrmware's most important mechanism for relay longevity. Relay contacts have a
ﬁnite switching lifetime, typically rated for 100,000 operations for a 10A resistive load. If
the ﬁrmware wrote to the relay pin on every motion event regardless of whether the state
had changed, a busy laboratory could generate thousands of unnecessary relay operations
per day. By comparing the commanded state to the last applied state and only toggling on
a genuine change, the ﬁrmware reduces relay operations to the minimum required.

The motionIgnored ﬂag allows the backend to acknowledge a sensor POST without
commanding any relay change --- for example, because the motion event occurred outside
a scheduled session and the administrator has conﬁgured the system to ignore after-hours
motion. When motionIgnored is true, lastToggleTime is updated to suppress further
motion events during the cooldown period, and the function returns without logging any
relay state.

Finally, wasNear is updated at the bottom of every loop() iteration (outside the
motion-event block) to track the previous presence state for the next rising-edge detection.



The delay(100) at the very end paces the loop at approximately 10 Hz, giving the DHT11
sensor sufﬁcient time between reads to avoid NaN responses while keeping the ultrasonic
presence detection responsive enough for human entry events.



\section{Python AI Processing Service Implementation}


The Python AI processing service is the second tier of the GreenLab architecture. It is
responsible for consuming a live video stream from a camera installed in each laboratory,
running the YOLO v11 object detection model on that stream, applying ByteTrack identity
tracking so that the same person is not counted multiple times across frames, and exposing
the resulting occupancy count via a FastAPI HTTP endpoint that the .NET Core backend
can poll.

The service was implemented in two stages. The ﬁrst stage --- a working prototype
in cout\_code.txt --- established the core architecture: a single shared YOLO model, one
background thread per camera, a threading.Event for readiness signalling, and a FastAPI
endpoint. The second stage --- the production service in droidip.py, supported by the
camera\_sources.py abstraction --- added multi-format camera input, ByteTrack tracking,
a typed CameraState dataclass, and a background cleanup worker for idle cameras. This
section presents all three ﬁles in their entirety, function by function.



\subsection{Stage 1: Single-Camera Prototype (cout\_code.txt)}


cout\_code.txt is 116 lines long and contains four functional components: import decla-
rations, a shared YOLO model loader, a per-camera detection loop, a thread-start helper,
and the FastAPI endpoint. Each is presented and explained in full below.


5.3.1.1 Import Declarations and Application Bootstrap

1 import threading
2 import time
3 from ultralytics import YOLO
4 import cv2
5 from fastapi import FastAPI, HTTPException




6 import uvicorn

8 app = FastAPI()



\textit{Listing 5.17: Module-level imports and FastAPI application object (cout\_code.txt, lines 1--8).}



Six imports cover the full dependency surface of the prototype. threading provides
Thread, Lock, and Event --- the three primitives used to make the model and camera state
safely shareable across concurrent requests. time provides time.sleep(), used in the
reconnection backoff. ultralytics provides the YOLO class, which encapsulates model
loading, preprocessing, inference, and post-processing in a single object. cv2 (OpenCV)
provides VideoCapture for camera access and resize for frame downscaling. fastapi
and HTTPException provide the REST endpoint infrastructure. uvicorn is the ASGI
server that hosts the FastAPI application when the script is run directly.


5.3.1.2 Shared YOLO Model --- Lazy Singleton with Double-Checked Locking

1 \# Shared model -- loaded ONCE
2 \_model = None
3 \_model\_lock = threading.Lock()

5 def get\_model():
6 global \_model
7 if \_model is None: \# First check (no lock -- fast path)
8 with \_model\_lock: \# Acquire lock only when needed
9 if \_model is None: \# Second check (inside lock -- safe)
10 \_model = YOLO("yolo11n.pt")
11 \_model.to("cpu")
12 return \_model


\textit{Listing 5.18: get\_model(): double-checked locking singleton for YOLO model (cout\_code.txt, lines}

11--20).


The double-checked locking pattern is used here for a speciﬁc reason: loading a
YOLO model from disk takes 2--5 seconds and allocates several hundred megabytes of
memory. If two camera threads start simultaneously (e.g., two requests arrive at the /current-count
endpoint before either thread has loaded the model), the naïve solution of locking and



checking only once inside the lock would still be correct --- both threads would acquire the
lock sequentially, one loads the model, the other ﬁnds \_model already set. The double-
checked pattern adds the outer if \_model is None check before acquiring the lock, so
that all subsequent calls after the ﬁrst return immediately without locking --- a signiﬁcant
performance gain for a lock-heavy concurrent server.

In Python, the GIL (Global Interpreter Lock) provides an additional safety guarantee:
the assignment \_model = YOLO(...) is atomic from the GIL's perspective, so the outer
check cannot observe a partially-initialised object. The double-checked pattern is therefore
both correct and efﬁcient in CPython.

\_model.to("cpu") explicitly places the model on the CPU rather than a GPU. This
is appropriate for a deployment on a laptop or server without a CUDA-capable GPU. If the
deployment hardware has a GPU, changing "cpu" to "cuda" would accelerate inference
by an order of magnitude, but yolo11n (the nano variant) is already fast enough on CPU
for the sampling cadence this service uses.


5.3.1.3 Camera State Registry

1 camera\_states = {}
2 camera\_lock = threading.Lock()


\textit{Listing 5.19: Global camera state dictionary and its protecting lock (cout\_code.txt, lines 22--23).}



camera\_states is a plain Python dictionary keyed by integer device index. Each
value is itself a dictionary with keys "count", "ready" (a threading.Event), and "thread"
(the running Thread object). camera\_lock is a threading.Lock that must be held when-
ever camera\_states is read or written from any thread, to prevent concurrent dictionary
modiﬁcations from corrupting its internal state --- Python dictionaries are not thread-safe
for concurrent reads and writes even though individual read operations are GIL-protected.


5.3.1.4 Per-Camera Detection Loop --- camera\_loop()

1 def camera\_loop(device\_index: int, ready\_event: threading.Event):
2 model = get\_model() \# shared reference -- no lock needed here

4 while True: \# outer reconnection loop




5 \# Primary open attempt: no backend hint
6 cap = cv2.VideoCapture(device\_index)
7 if not cap.isOpened():
8 cap.release()
9 \# Fallback: DirectShow backend (Windows virtual cameras)
10 cap = cv2.VideoCapture(device\_index, cv2.CAP\_DSHOW)


12 \# Tune buffer to reduce latency for virtual cameras
13 cap.set(cv2.CAP\_PROP\_BUFFERSIZE, 2)
14 cap.set(cv2.CAP\_PROP\_FRAME\_WIDTH, 640)
15 cap.set(cv2.CAP\_PROP\_FRAME\_HEIGHT, 480)


17 if not cap.isOpened():
18 print(f"[ERROR] Cannot open camera index {device\_index}, retrying in 3s
...")
19 time.sleep(3)
20 continue \# retry outer loop

22 print(f"[INFO] Connected to camera index: {device\_index}")


\textit{Listing 5.20: camera\_loop() Part A: reconnection loop, two-attempt open strategy, and buffer}

conﬁguration (cout\_code.txt, lines 26--50).


camera\_loop() calls get\_model() at the top, which returns the already-loaded shared
model on all subsequent calls after the ﬁrst. The returned model object is used in the inner
loop for every inference call without any additional locking, which is safe because YOLO
model inference is a read-only operation on the model's weights --- the model object is
never modiﬁed during inference.

The two-attempt camera open strategy addresses a speciﬁc quirk of virtual camera
software on Windows: DroidCam (which turns an Android phone into a USB webcam) and
its alternatives sometimes only enumerate correctly with the DirectShow backend (cv2.CAP\_DSHOW)
rather than OpenCV's default backend. The primary attempt without a backend hint works
on Linux and macOS; the CAP\_DSHOW fallback is the Windows safety net. Calling cap.release()
between attempts is essential to free the ﬁle descriptor before attempting a new open.

Setting CAP\_PROP\_BUFFERSIZE to 2 reduces the number of frames OpenCV buffers
internally from the default (typically 5--10 for USB cameras) to 2. A smaller buffer means



cap.read() returns a more recent frame --- important for occupancy detection where we
want the current count, not the count from three seconds ago. At 640×480 and 30 fps, two
buffered frames represent approximately 67 ms of latency, which is acceptable.

1 while True: \# inner frame-reading loop
2 ret, frame = cap.read()
3 if not ret:
4 print(f"[WARN] Lost connection on cam {device\_index}, reconnecting
...")
5 break \# break inner reconnect outer ->

7 \# Downsample to 320x320 for faster inference
8 small\_frame = cv2.resize(frame, (320, 320))
9 results = model(small\_frame, verbose=False, imgsz=320)

11 \# Count class-0 (person) detections above confidence threshold
12 person\_count = sum(
13 1
14 for r in results
15 if r.boxes is not None
16 for box in r.boxes
17 if int(box.cls[0]) == 0 and float(box.conf[0]) > 0.5
18 )

20 with camera\_lock:
21 camera\_states[device\_index]["count"] = person\_count


23 \# Signal ready after first successful detection pass
24 if not ready\_event.is\_set():
25 ready\_event.set()


27 cap.release()
28 time.sleep(1) \# brief pause before reconnect


\textit{Listing 5.21: camera\_loop() Part B: frame reading, YOLO inference, person counting, and state}

update (cout\_code.txt, lines 51--76).


The inner frame loop reads frames continuously and runs inference on every frame.
cv2.resize(frame, (320, 320)) downsamples the 640×480 frame (or whatever the



camera delivers) to 320×320 pixels before inference. This is a deliberate accuracy/speed
trade-off: yolo11n was trained on 640×640 images and performs best at its native resolu-
tion, but at 320×320 it still reliably detects persons at distances of 1--5 metres (typical for a
laboratory doorway sensor) while running approximately four times faster than at 640×640
on CPU.

The person counting generator expression is a nested comprehension that iterates over
all Results objects returned by model() (one per image, so effectively one object here),
then over all detected bounding boxes in that result, and yields 1 for each box where the
\begin{lstlisting}
class index (box.cls[0]) is 0 (the COCO class index for ``person'') and the conﬁdence
\end{lstlisting}

score (box.conf[0]) exceeds 0.5. The sum() of those 1s is the frame-level person count.

The ready\_event.set() call signals the FastAPI endpoint that at least one frame
has been successfully processed. The endpoint blocks on ready\_event.wait() until this
signal is received, preventing it from returning a stale zero count on the very ﬁrst request
to a freshly started camera. The is\_set() check before calling set() avoids a redundant
syscall on every frame after the ﬁrst.


5.3.1.5 Thread-Start Helper --- \_start\_camera()

1 def \_start\_camera(cam\_id: int):
2 """Start camera thread only if not already running.
3 Must be called with camera\_lock held."""
4 if cam\_id not in camera\_states:
5 ready\_event = threading.Event()
6 camera\_states[cam\_id] = {"count": 0,
7 "ready": ready\_event,
8 "thread": None}
9 t = threading.Thread(
10 target=camera\_loop,
11 args=(cam\_id, ready\_event),
12 daemon=True
13 )
14 t.start()
15 camera\_states[cam\_id]["thread"] = t


\textit{Listing 5.22: \_start\_camera(): guarded thread creation, must be called with camera\_lock held}

(cout\_code.txt, lines 79--93).



\_start\_camera() is a private helper that must be called with camera\_lock already
held --- its docstring makes this contract explicit. The cam\_id not in camera\_states
guard prevents starting a second thread for a camera that is already running, which is
the primary race condition this helper needs to prevent: if two HTTP requests for the
same cam\_id arrive simultaneously, both will pass the outer get\_count() check, but only
the ﬁrst to acquire camera\_lock and call \_start\_camera() will ﬁnd cam\_id not in
camera\_states; the second will ﬁnd it already present and do nothing.

Setting daemon=True on the thread means Python will not wait for the camera loop
to ﬁnish before the main process exits. This is the correct choice for a background service
thread: if the FastAPI server shuts down (SIGTERM, Ctrl+C), the camera threads should
not block the shutdown process.



5.3.1.6 FastAPI Endpoint --- /current-count

1 @app.get("/current-count")
2 def get\_count(cam\_id: int = 0):
3 """
4 Pass the camera index.
5 Usually 0 is your built-in webcam, 1 is DroidCam/DropCam USB.
6 """
7 with camera\_lock:
8 if cam\_id not in camera\_states:
9 \_start\_camera(cam\_id) \# Safe: lock is already held
10 ready\_event = camera\_states[cam\_id]["ready"]


12 \# Wait OUTSIDE the lock -- do not hold it for up to 10 seconds
13 if not ready\_event.wait(timeout=10):
14 raise HTTPException(
15 status\_code=503,
16 detail=f"Camera {cam\_id} not responding."
17 f" Check your DropCam/DroidCam connection.",
18 )

20 with camera\_lock:
21 state = camera\_states.get(cam\_id, {"count": 0})






23 return {"people\_count": state["count"], "camera\_index": cam\_id}



26 if \_\_name\_\_ == "\_\_main\_\_":
27 uvicorn.run(app, host="0.0.0.0", port=8000)


\textit{Listing 5.23: /current-count endpoint: lazy thread start, ready-event blocking (10 s timeout), and}

count retrieval (cout\_code.txt, lines 95--116).


The endpoint acquires camera\_lock, checks whether the requested cam\_id has a
running thread, and if not calls \_start\_camera() (which expects the lock to be held). It
then extracts the ready\_event reference from the state dictionary and immediately releases
the lock. The release is critical: ready\_event.wait(timeout=10) may block for up
to 10 seconds, and holding camera\_lock for 10 seconds would prevent any other HTTP
request from being processed during that time --- a deadlock-adjacent scenario that would
make the service effectively single-threaded under load.

ready\_event.wait() returns True if the event was set within the timeout, False if
the timeout expired without the event being set. A False return means the camera thread
started but did not produce a single successful frame within 10 seconds --- either the camera
is not connected, DroidCam is not running on the phone, or the camera index is wrong.
The 503 Service Unavailable response with a human-readable detail message allows the
backend to present this as a camera fault rather than an application error.

The second camera\_lock acquisition for the ﬁnal count read is a separate critical
section from the ﬁrst. Between the end of the ﬁrst critical section and the start of the second,
the camera thread may have updated camera\_states[cam\_id]["count"] any number of
times --- this is acceptable, as the endpoint simply returns whatever count is current at the
moment it reads it.



\subsection{Camera Source Abstraction Module (camera\_sources.py)}


camera\_sources.py is 189 lines long. It is a pure-utility module with no FastAPI depen-
dency and no global state --- it consists entirely of functions that translate camera source
strings into values cv2.VideoCapture can open. It was extracted from droidip.py to
make the source-resolution logic independently testable and reusable. The module deﬁnes



one regex constant, one frozenset constant, three public functions, and two private helper
functions.


5.3.2.1 Module-Level Constants

1 import logging
2 import os
3 import re
4 from typing import Union

6 import cv2

8 logger = logging.getLogger("people\_counter.camera\_sources")

10 \# Supported video file extensions
11 \_VIDEO\_EXTENSIONS: frozenset[str] = frozenset({
12 ".mp4", ".avi", ".mov", ".mkv", ".flv", ".ts"
13 })

15 \# Pattern that matches a bare IPv4 address (no scheme, no path)
16 \_BARE\_IP\_RE = re.compile(
17 r"^(\d{1,3}\.){3}\d{1,3}$"
18 )


\textit{Listing 5.24: camera\_sources.py: imports, logger, supported ﬁle extensions frozenset, and bare-IP}

regex (camera\_sources.py, lines 1--17).


\_VIDEO\_EXTENSIONS is a frozenset (immutable set) of lowercase ﬁle extensions. Us-
ing frozenset rather than a list makes membership testing O(1) rather than O(n), which
matters when resolve\_source() is called on every incoming API request. The extensions
cover the most common container formats that OpenCV can decode: .mp4 (H.264/H.265
in MP4), .avi (various codecs in AVI), .mov (QuickTime), .mkv (Matroska), .flv (Flash
Video), and .ts (MPEG Transport Stream).

\_BARE\_IP\_RE is a compiled regular expression that matches exactly a dotted-quad
IPv4 address with no scheme, port, or path. The pattern requires: start of string, three
groups of 1--3 digits followed by a literal dot, one ﬁnal group of 1--3 digits, end of string. It
does not validate that each octet is in the range 0--255 --- 999.999.999.999 would match



--- but for a camera IP address entered by an administrator this is an acceptable trade-off. A
stricter regex would be longer and harder to read without providing meaningful additional
protection, since an invalid IP would simply fail to connect.



5.3.2.2 resolve\_source() --- The Core Dispatch Function

1 def resolve\_source(source: str, default\_port: int = 4747) -> Union[str, int]:
2 """
3 Translate a caller-supplied source string into the value that
4 cv2.VideoCapture expects: an integer for USB devices, a string
5 for everything else.

7 Parameters
8 ----------
9 source : Raw source string from the API caller.
10 default\_port : Port appended when source is a bare IP address.

12 Returns
13 -------
14 int for USB / device-index sources. str for network streams and file paths.

16 Raises
17 ------
18 ValueError if the source format is not recognised.
19 """
20 source = source.strip()
21 if not source:
22 raise ValueError("Camera source must not be empty")

24 if source.lower().startswith("usb:"):
25 return \_parse\_usb(source)


27 if source.startswith(("http://", "https://", "rtsp://")):
28 logger.debug("Network stream source: %s", source)
29 return source


31 ext = os.path.splitext(source)[1].lower()
32 if ext in \_VIDEO\_EXTENSIONS:




33 return \_parse\_video\_file(source)

35 if \_BARE\_IP\_RE.match(source):
36 url = f"http://{source}:{default\_port}/video"
37 logger.debug("Bare IP expanded to: %s", url)
38 return url


40 raise ValueError(
41 f"Unrecognised camera source: '{source}'. "
42 "Expected usb:<index>, http/https/rtsp URL, "
43 "a video file (.mp4/.avi/.mov), or a bare IPv4 address."
44 )


\textit{Listing 5.25: resolve\_source(): ﬁve-branch dispatch with full docstring (camera\_sources.py, lines}

20--70).


resolve\_source() applies its ﬁve recognition patterns in a strict priority order that
prevents ambiguity: the usb: preﬁx check comes ﬁrst, before any other pattern, so that a
hypothetically unusual string like "usb:192.168.1.1" is handled as a USB source request
rather than being misidentiﬁed as an IP address. Network scheme checks come second;
these are the most speciﬁc and most common runtime inputs. The video ﬁle extension
check comes third; it only ﬁres if the source does not have a network scheme but does have
a recognised extension. The bare-IP regex comes fourth; it is the backward-compatibility
catch-all for DroidCam-style inputs. The ValueError at the end is the catch-all for inputs
that match none of the above.

The source.strip() at the top removes leading and trailing whitespace, guarding
against inputs like " 192.168.1.3 " pasted from a text ﬁeld with accidental spaces. The
empty-string check after strip() catches inputs that were whitespace-only --- without it,
a whitespace-only string would pass the strip() and then fall through all ﬁve pattern
checks to the ValueError, but the error message would say ``Unrecognised camera source:
'' '' rather than the more informative ``Camera source must not be empty''.


5.3.2.3 create\_capture() --- Opening the VideoCapture

1 def create\_capture(source: str, default\_port: int = 4747) -> cv2.VideoCapture:
2 """




Table 5.3: Example inputs to resolve\_source() and their resolved output.

Input example Branch taken Output
usb:0 USB preﬁx int: 0
usb:2 USB preﬁx int: 2
Network
http://192.168.1.3:4747/video str: unchanged
scheme
Network
rtsp://admin:pass@10.0.0.5/live str: unchanged
scheme
/recordings/lab\_d7.mp4File extension str: unchanged
(.mp4)
192.168.1.3 Bare IP regex str:
http://192.168.1.3:4747/video
192.168.1.3 None (Val- ValueError: Unrecognised
port=8080 ueError) source
ftp://camera/stream None (Val- ValueError: Unrecognised
ueError) source


3 Parse source and return an opened cv2.VideoCapture.
4 Callers should check cap.isOpened() themselves.
5 """
6 resolved = resolve\_source(source, default\_port)

8 if isinstance(resolved, int):
9 logger.info("Opening USB device index %d", resolved)
10 cap = cv2.VideoCapture(resolved)
11 else:
12 logger.info("Opening stream: %s", resolved)
13 cap = cv2.VideoCapture(resolved)


15 if not cap.isOpened():
16 logger.warning(
17 "VideoCapture did not open immediately for source: %s", source
18 )

20 return cap


\textit{Listing 5.26: create\_capture(): resolves source and returns VideoCapture without asserting success}

(camera\_sources.py, lines 72--100).


create\_capture() intentionally does not raise an exception when the VideoCapture
fails to open. The reason is that different callers have different retry and logging strate-



gies: the production camera loop in droidip.py logs a warning and sleeps 3 seconds
before retrying; a test harness might want to assert immediately. By returning the unopened
VideoCapture object and leaving the isOpened() check to the caller, create\_capture()
is maximally composable. The warning log inside create\_capture() provides visibility
without forcing a particular error-handling strategy.

The isinstance(resolved, int) branch exists because cv2.VideoCapture(0)
and cv2.VideoCapture("http://...") call the same constructor with different types.
Although Python would handle both correctly without the branch (since VideoCapture
accepts both int and str), the explicit branch allows the logger to emit a more informative
message --- ``Opening USB device index 0'' versus ``Opening stream: http://...'' --- which is
valuable for diagnosing connectivity problems in a deployed system.


5.3.2.4 source\_label() --- Stable Registry Key Generation

1 def source\_label(source: str, default\_port: int = 4747) -> str:
2 """
3 Return a human-readable, stable label for source suitable for use
4 as a registry key. For network sources this is the fully-expanded
5 URL; for USB and file sources it is the canonical string.
6 """
7 try:
8 resolved = resolve\_source(source, default\_port)
9 return str(resolved) \# int -> "0", str -> unchanged
10 except ValueError:
11 return source \# fall back to raw string if unresolvable


\textit{Listing 5.27: source\_label(): stable key generation for the camera registry (camera\_sources.py,}

lines 102--115).


source\_label() is used by droidip.py as the key for its \_cameras registry dictio-
nary. Using the fully-resolved URL as the key ensures that 192.168.1.3 and http://192.168.1.3:4747
--- two different input strings that refer to the same physical camera --- resolve to the same
key, preventing accidental double-registration of the same camera under two different input
formats. str(resolved) converts int device indices like 0 to the string "0", so all registry
keys are strings regardless of input type. The except ValueError fallback returns the
raw source string, which would then be stored as-is in the registry --- a pragmatic choice



that allows the registry to absorb inputs that will ultimately fail at the VideoCapture level
without crashing the key-generation step.


5.3.2.5 Private Helpers --- \_parse\_usb() and \_parse\_video\_ﬁle()

1 def \_parse\_usb(source: str) -> int:
2 """Parse 'usb:<index>' and return the integer device index."""
3 \_, \_, index\_str = source.partition(":")
4 index\_str = index\_str.strip()


6 if not index\_str.lstrip("-").isdigit():
7 raise ValueError(
8 f"Invalid USB source '{source}': index must be a non-negative integer "
9 f"(e.g. usb:0, usb:1)."
10 )

12 index = int(index\_str)
13 if index < 0:
14 raise ValueError(
15 f"Invalid USB source '{source}': device index must be >= 0, got {index}.
"
16 )


18 logger.debug("USB device index: %d", index)
19 return index


22 def \_parse\_video\_file(source: str) -> str:
23 """
24 Validate that source is a readable file path and return it unchanged.
25 Only warns (does not raise) when the file does not exist.
26 """
27 if not os.path.isfile(source):
28 logger.warning(
29 "Video file not found at '%s'. "
30 "Make sure the path is accessible from the server's working directory.",
31 source,
32 )
33 else:




34 logger.debug("Video file source: %s", source)

36 return source


\textit{Listing 5.28:     \_parse\_usb() and \_parse\_video\_ﬁle():       type-speciﬁc validation helpers}

(camera\_sources.py, lines 117--189).


\_parse\_usb() uses str.partition(":") rather than str.split(":") to correctly
handle hypothetical inputs like "usb:0:extra" --- partition always returns exactly three
parts (before the ﬁrst delimiter, the delimiter, and everything after), so index\_str would be
"0:extra" in that case, which would then fail the isdigit() check and raise a ValueError
with a clear message. The lstrip("-") before isdigit() allows the isdigit() check to
see past a leading minus sign --- so "usb:-1" would pass isdigit() on "-1".lstrip("-")
= "1", but then fail the index < 0 check with a more speciﬁc error message than a generic
isdigit() failure.

\_parse\_video\_file() does not raise on a missing ﬁle by design: not raising, and
just logging a warning, reﬂects the fact that the working directory at validation time may
differ from the working directory at inference time. A Docker container, for example,
might mount a recordings volume at /data/recordings at runtime, which does not exist
on the host ﬁlesystem where validation runs. Treating a missing ﬁle as a warning rather
than an error allows the service to accept a ﬁle path conﬁguration even before the volume
is mounted.



\subsection{Stage 2: Production Multi-Camera Tracking API (droidip.py)}


droidip.py is 322 lines long and is the service that the Chapter 3 architecture diagrams
describe as the AI Processing tier. It builds on every pattern established in the prototype but
replaces integer device-index inputs with source-string resolution, replaces raw per-frame
box counting with ByteTrack-based identity tracking, replaces the loose dictionary state
with a typed CameraState dataclass, and adds a background cleanup worker that removes
idle cameras from the registry.


5.3.3.1 Module Docstring and Conﬁguration Constants





1 """
2 People Counter API
3 ==================
4 FastAPI service that streams YOLO-tracked person counts from IP cameras.

6 Supported stream formats:
7 - DroidCam: /current-count?ip=192.168.1.3
8 - HTTP: /current-count?ip=http://192.168.1.3:4747/video
9 - RTSP: /current-count?ip=rtsp://admin:pass@camera/live
10 - IP+port: /current-count?ip=192.168.1.3\&port=8080
11 """

13 MODEL\_PATH: str = "yolo11n.pt"
14 MODEL\_DEVICE: str = "cpu"
15 CONFIDENCE\_THRESHOLD: float = 0.5
16 FRAME\_SKIP: int = 3
17 READY\_TIMEOUT: float = 15.0
18 CAMERA\_TIMEOUT: float = 300.0
19 CLEANUP\_INTERVAL: float = 60.0
20 DEFAULT\_PORT: int = 4747



\textit{Listing 5.29: Module docstring and all conﬁguration constants (droidip.py, lines 1--28).}




The module docstring lists all four supported input formats for the ip query parameter,
making the service's capabilities self-documenting without requiring a separate README.
Each constant is annotated with its type for static analysis tool compatibility. The values re-
ﬂect the operational requirements: CONFIDENCE\_THRESHOLD = 0.5 satisﬁes FR-OD-01's
minimum detection conﬁdence; FRAME\_SKIP = 3 balances inference frequency against
CPU load; READY\_TIMEOUT = 15.0 gives a freshly-started camera 15 seconds to deliver
its ﬁrst frame before returning a 503 (extended from the prototype's 10 seconds to account
for RTSP cameras that are slower to negotiate stream parameters); CAMERA\_TIMEOUT =
\section{allows a camera that has not been polled for ﬁve minutes to be garbage-collected;}

CLEANUP\_INTERVAL = 60.0 runs the cleanup worker every minute; DEFAULT\_PORT =
4747 is the DroidCam default port.






5.3.3.2 Structured Logging Conﬁguration

1 import logging

3 logging.basicConfig(
4 level=logging.INFO,
5 format="%(asctime)s [%(levelname)s] %(name)s: %(message)s",
6 datefmt="%Y-%m-%d %H:%M:%S",
7 )
8 logger = logging.getLogger("people\_counter")


\textit{Listing 5.30: Structured logging setup with timestamp, level, and logger-name format (droidip.py,}

lines 30--40).


The logging conﬁguration adds a timestamp, log level, and logger name to every mes-
sage. This is a signiﬁcant improvement over the prototype's print() statements: print()
output is unstructured and interleaved with other output in a multi-threaded environment;
logging is thread-safe (Python's logging module uses an internal lock) and each mes-
sage includes enough context to diagnose multi-camera scenarios where several threads
are producing output simultaneously. The format string produces output like ``2026-06-04
09:15:32 [INFO] people\_counter: Connected to stream: http://192.168.1.3:4747/vide
which is immediately grep-able by stream URL or log level in a deployed environment.

camera\_sources.py uses a child logger named "people\_counter.camera\_sources",
which is a child of the "people\_counter" logger conﬁgured here. Python's logging hier-
archy means that messages from the child logger propagate to the parent by default, so a
single basicConfig() call covers all modules in the service.


5.3.3.3 YOLO Model Singleton --- Production Version

1 \_model: Optional[YOLO] = None
2 \_model\_lock = threading.Lock()

4 def get\_model() -> YOLO:
5 """Return the shared YOLO model, loading it on first call."""
6 global \_model
7 if \_model is None:
8 with \_model\_lock:




9 if \_model is None:
10 logger.info("Loading YOLO model: %s on %s", MODEL\_PATH, MODEL\_DEVICE
)
11 \_model = YOLO(MODEL\_PATH)
12 \_model.to(MODEL\_DEVICE)
13 logger.info("Model loaded successfully")
14 return \_model


\textit{Listing 5.31: Production get\_model(): same double-checked locking pattern as the prototype, with}

structured logging (droidip.py, lines 48--62).


The production get\_model() is structurally identical to the prototype version but re-
places print() with logger.info() and uses the MODEL\_PATH and MODEL\_DEVICE con-
stants rather than hard-coded strings. The Optional[YOLO] type annotation on \_model
enables static type checkers to verify that all code paths that use \_model account for the
possibility that it is None before the ﬁrst call to get\_model().


5.3.3.4 CameraState Dataclass and Registry

1 from dataclasses import dataclass, field

3 @dataclass
4 class CameraState:
5 count: int = 0
6 connected: bool = False
7 last\_access: float = field(default\_factory=time.monotonic)
8 ready\_event: threading.Event = field(default\_factory=threading.Event)


10 \# registry: stream\_url -> CameraState
11 \_cameras: dict[str, CameraState] = {}
12 \_cameras\_lock = threading.Lock()


\textit{Listing 5.32: CameraState dataclass with ﬁeld() defaults and the camera registry (droidip.py, lines}

64--78).


Replacing the prototype's loose dictionary ({"count": 0, "ready": event, "thread":
t}) with a typed dataclass provides three beneﬁts. First, ﬁeld names are validated at
\begin{lstlisting}
class deﬁnition time rather than at runtime --- a typo like state["connt"] would be



\end{lstlisting}

caught by a type checker or raise a KeyError immediately, whereas setattr(state,
"connt", value) in the production \_update\_state() helper would silently create a
new attribute. Second, the @dataclass decorator auto-generates \_\_init\_\_, \_\_repr\_\_,
and \_\_eq\_\_ methods, which are useful for testing and debugging. Third, the field()
function with default\_factory=time.monotonic means that last\_access is set to the
current monotonic clock value at the moment the CameraState is instantiated --- if a sim-
ple default value like 0.0 were used instead, all newly-created cameras would have the
same last\_access, potentially causing the cleanup worker to remove them immediately
after creation if cleanup runs before the ﬁrst request.

time.monotonic() is used rather than time.time() because monotonic() is guar-
anteed never to go backwards --- system clock adjustments (NTP sync, daylight saving
time transitions) cannot cause last\_access to jump forward and falsely make a recently-
accessed camera appear stale.


5.3.3.5 Thread-Safe State Update Helper --- \_update\_state()

1 def \_update\_state(url: str, **kwargs) -> None:
2 """Thread-safe helper to patch fields on an existing CameraState."""
3 with \_cameras\_lock:
4 state = \_cameras.get(url)
5 if state is None:
6 return \# camera was removed by cleanup -- silently ignore
7 for key, value in kwargs.items():
8 setattr(state, key, value)


\textit{Listing 5.33: \_update\_state(): keyword-argument-based ﬁeld patcher with cleanup-race guard}

(droidip.py, lines 80--90).


\_update\_state() is a convenience function that takes keyword arguments and ap-
plies them to the CameraState for a given URL, all within a single lock acquisition. With-
out this helper, every state update in the camera loop would require three lines: acquire lock,
set ﬁeld, release lock. The **kwargs variadic signature allows updating multiple ﬁelds
in a single call: \_update\_state(url, connected=False) or \_update\_state(url,
connected=True, count=0). The state is None guard handles the race condition
where the cleanup worker removes the camera from the registry between the camera loop's



registration check and the update call --- rather than raising a KeyError or AttributeError,
it silently returns, allowing the camera loop to detect the deregistration on its next iteration.


5.3.3.6 Production Camera Worker Loop --- \_camera\_loop()

1 def \_camera\_loop(stream\_url: str) -> None:
2 """
3 Long-running thread: opens the stream, runs YOLO ByteTrack inference,
4 and updates the shared CameraState. Reconnects on failure.
5 """
6 model = get\_model()

8 while True: \# outer reconnection loop
9 \# Check deregistration before each reconnect attempt
10 with \_cameras\_lock:
11 if stream\_url not in \_cameras:
12 logger.info("Camera deregistered, worker exiting: %s", stream\_url)
13 return

15 cap = create\_capture(stream\_url)
16 if not cap.isOpened():
17 logger.warning("Cannot open stream, retrying in 3 s: %s", stream\_url)
18 \_update\_state(stream\_url, connected=False)
19 time.sleep(3)
20 continue

22 logger.info("Connected to stream: %s", stream\_url)
23 \_update\_state(stream\_url, connected=True)
24 frame\_idx = 0


\textit{Listing 5.34: \_camera\_loop() Part A: outer reconnection loop with deregistration check (droidip.py,}

lines 93--120).


The outer reconnection loop begins with a deregistration check: if the cleanup worker
has removed stream\_url from \_cameras between the camera thread starting and it at-
tempting to open the stream, the thread exits cleanly without attempting a connection. This
is important for the correctness of the cleanup mechanism --- without this check, a dereg-
istered camera's thread would continue running indeﬁnitely, consuming CPU and camera



resources even though no endpoint is polling its count.

create\_capture() from camera\_sources.py abstracts the actual VideoCapture
construction. The camera loop does not need to know whether stream\_url is a USB index,
an RTSP URL, or a DroidCam HTTP stream --- create\_capture() handles all cases. The
connected=False state update before the 3-second sleep allows the /current-count
endpoint to report ``connected: false'' in its response, giving the caller a machine-readable
indicator that the camera is temporarily unavailable.

1 while True: \# inner frame-reading loop
2 \# Check deregistration before each frame read
3 with \_cameras\_lock:
4 if stream\_url not in \_cameras:
5 logger.info("Camera deregistered mid-loop, exiting: %s",
stream\_url)
6 cap.release()
7 return

9 ret, frame = cap.read()
10 if not ret:
11 logger.warning("Stream lost, reconnecting: %s", stream\_url)
12 \_update\_state(stream\_url, connected=False)
13 break \# break inner -> reconnect outer

15 frame\_idx += 1
16 if frame\_idx % FRAME\_SKIP != 0:
17 continue \# discard this frame without inference


19 try:
20 people\_count = \_run\_inference(model, frame)
21 except Exception:
22 logger.exception("Inference error on stream: %s", stream\_url)
23 continue

25 with \_cameras\_lock:
26 state = \_cameras.get(stream\_url)
27 if state is None:
28 cap.release()
29 return




30 state.count = people\_count
31 state.ready\_event.set()


33 cap.release()
34 time.sleep(1)


\textit{Listing 5.35: \_camera\_loop() Part B: inner frame loop with FRAME\_SKIP, exception guard, and}

deregistration-aware state update (droidip.py, lines 121--160).


The inner frame loop has a second deregistration check, inside the loop itself be-
fore each frame read. This ensures that a camera deregistered while the loop is run-
ning causes the thread to exit promptly --- without this check, the thread might read and
process hundreds of frames between cleanup events before the outer loop's check runs.
cap.release() is called before returning to free the VideoCapture's ﬁle descriptor and
any OS resources associated with the video stream.

FRAME\_SKIP = 3 means inference runs on frames 3, 6, 9, 12, . . . --- every third frame.
At a typical USB camera or DroidCam frame rate of 30 fps, this produces 10 inference runs
per second. At a slower 15 fps (some RTSP cameras), this produces 5 inference runs per
second. Both are well above the effective 0.2 runs per second that a ﬁve-second sampling
interval would imply. The frame\_idx counter is a local variable initialised to 0 when the
inner loop starts (after a successful open); it resets to 0 if the connection drops and the outer
loop reconnects, which is correct because the skipped-frame count should restart from a
fresh connection.

The try/except around \_run\_inference() prevents a single frame's inference er-
ror (for example, a malformed frame from a corrupted RTSP packet) from crashing the
entire camera thread. logger.exception() logs the full stack trace at ERROR level,
which is critical for post-hoc diagnosis of intermittent failures. The continue after the
exception leaves state.count unchanged, so the last successful count remains visible to
the endpoint.


5.3.3.7 ByteTrack Inference --- \_run\_inference()

1 def \_run\_inference(model: YOLO, frame) -> int:
2 """
3 Run ByteTrack inference on a single frame and return the number




4 of unique active tracked persons.
5 """
6 results = model.track(
7 frame,
8 persist=True, \# maintain track state across calls
9 tracker="bytetrack.yaml",
10 imgsz=320,
11 conf=CONFIDENCE\_THRESHOLD,
12 classes=[0], \# detect only class 0 = person
13 verbose=False,
14 )


16 tracked\_ids: set[int] = set()
17 for r in results:
18 if r.boxes is None:
19 continue
20 for box in r.boxes:
21 cls = int(box.cls[0])
22 conf = float(box.conf[0])
23 track\_id = box.id
24 if cls == 0 and conf >= CONFIDENCE\_THRESHOLD and track\_id is not None:
25 tracked\_ids.add(int(track\_id[0]))

27 return len(tracked\_ids)


\textit{Listing 5.36: \_run\_inference(): ByteTrack-based person tracking returning unique active track}

count (droidip.py, lines 163--195).



The key change from the prototype's model() call to the production model.track()
call is the addition of three arguments: persist=True, tracker="bytetrack.yaml", and
classes=[0]. persist=True tells YOLO to maintain the ByteTrack state between con-
secutive calls on the same camera stream --- without it, track IDs would be re-assigned from
zero on every call, making cross-frame identity impossible. tracker="bytetrack.yaml"
selects the ByteTrack algorithm from YOLO's bundled tracker conﬁgurations. classes=[0]
restricts detection to person only at the model level, before any post-processing, which re-
duces the number of boxes the tracker needs to manage and speeds up inference.

ByteTrack (Zhang et al., 2022) is a multi-object tracker that associates detection boxes



across frames using a combination of IoU (Intersection over Union) matching and Kalman
ﬁlter position prediction. It maintains a set of ``active'' tracks (conﬁrmed persons) and a set
of ``lost'' tracks (persons that have left the frame or been momentarily occluded). A track
is kept in the active set if it was successfully matched to a detection in the current frame;
it moves to the lost set if unmatched for one frame and is deleted if still unmatched after a
conﬁgured number of frames. persist=True ensures the lost-track buffer persists between
calls, so a person who bends down and brieﬂy disappears from the YOLO detection is not
immediately removed from the active set.

The tracked\_ids set collects the integer track ID of every person detection in the
frame that passes both the conﬁdence threshold and the track\_id is not None check.
The track\_id is not None check is necessary because model.track() may return boxes
without track IDs in edge cases (e.g., the ﬁrst frame before ByteTrack has initialised its
state). len(tracked\_ids) is the count of unique person identities currently visible in the
frame --- the number the backend will use as occupancyCount.


5.3.3.8 Camera Lifecycle Management --- \_ensure\_camera\_started() and \_cleanup\_worker()

1 def \_ensure\_camera\_started(stream\_url: str) -> CameraState:
2 """
3 Return the CameraState for stream\_url, creating and starting a
4 worker thread if this is the first time we see this URL.
5 """
6 with \_cameras\_lock:
7 if stream\_url not in \_cameras:
8 logger.info("Registering new camera: %s", stream\_url)
9 state = CameraState()
10 \_cameras[stream\_url] = state

12 worker = threading.Thread(
13 target=\_camera\_loop,
14 args=(stream\_url,),
15 daemon=True,
16 name=f"cam-{stream\_url[-30:]}",
17 )
18 worker.start()
19 else:




20 state = \_cameras[stream\_url]
21 state.last\_access = time.monotonic()


23 return state


\textit{Listing 5.37: \_ensure\_camera\_started(): thread creation on ﬁrst request, last\_access refresh on}

subsequent requests (droidip.py, lines 198--220).


\_ensure\_camera\_started() is called by the /current-count endpoint on every
request. The entire registration-and-thread-start sequence runs inside a single camera\_lock
acquisition, which atomically checks whether the camera is already registered and either
creates a new CameraState and starts a thread, or updates last\_access on the existing
state. The name= argument on the Thread object sets the thread's name to the last 30
characters of the stream URL, which makes thread names visible in Python proﬁlers and
debuggers --- far more informative than the default ``Thread-1'', ``Thread-2'' naming.

state.last\_access = time.monotonic() on the existing-camera path is the heart-
beat mechanism for the cleanup worker. Every poll of a camera resets its expiry timer. A
camera that is polled every few seconds will never become stale; a camera whose poll stops
(e.g., the backend goes ofﬂine) will have its last\_access frozen at the time of the last poll
and will eventually be cleaned up after CAMERA\_TIMEOUT seconds.

1 def \_cleanup\_worker() -> None:
2 """Background thread: remove cameras idle for > CAMERA\_TIMEOUT seconds."""
3 while True:
4 time.sleep(CLEANUP\_INTERVAL)
5 cutoff = time.monotonic() - CAMERA\_TIMEOUT

7 with \_cameras\_lock:
8 stale = [
9 url for url, s in \_cameras.items()
10 if s.last\_access < cutoff
11 ]
12 for url in stale:
13 logger.info(
14 "Removing inactive camera (idle > %.0f s): %s",
15 CAMERA\_TIMEOUT, url
16 )




17 del \_cameras[url]

19 \# Start the cleanup worker when the module loads
20 \_cleanup\_thread = threading.Thread(
21 target=\_cleanup\_worker,
22 daemon=True,
23 name="cleanup-worker",
24 )
25 \_cleanup\_thread.start()


\textit{Listing 5.38: \_cleanup\_worker(): idle-camera garbage collection and module-level thread start}

(droidip.py, lines 223--248).


\_cleanup\_worker() runs on a daemon thread started at module import time. It
sleeps for CLEANUP\_INTERVAL (60) seconds, then calculates a cutoff timestamp as the cur-
rent monotonic time minus CAMERA\_TIMEOUT (300 seconds). Any camera whose last\_access
is older than this cutoff is considered stale. The stale list is built with a list comprehension
inside the lock (safe --- iterating over a dictionary while holding its lock), and then the
stale URLs are deleted inside the same lock acquisition. Deleting from the dictionary in-
side the lock is the operation that the camera loop's deregistration checks look for: after del
\_cameras[url], the next time the camera thread checks stream\_url not in \_cameras,
it will return True and the thread will exit.

Building the stale list separately from deleting prevents a ``dictionary changed size
during iteration'' RuntimeError, which would occur if del \_cameras[url] were called
inside the for url, s in \_cameras.items() iteration without a separate list.


5.3.3.9 FastAPI Endpoints --- /current-count, /health, and /

1 app = FastAPI(title="People Counter API", version="2.0.0")

3 @app.get("/current-count")
4 def get\_count(ip: str, port: int = DEFAULT\_PORT):
5 """
6 Return the current tracked-person count for a camera.


8 ip : Camera IP or full URL (http/https/rtsp).
9 port : Port for bare-IP sources (default: 4747).




10 """
11 if not ip:
12 raise HTTPException(status\_code=400, detail="ip parameter is required")

14 stream\_url = build\_stream\_url(ip, port)
15 state = \_ensure\_camera\_started(stream\_url)


17 if not state.ready\_event.wait(timeout=READY\_TIMEOUT):
18 raise HTTPException(
19 status\_code=503,
20 detail=f"Camera stream not ready within {READY\_TIMEOUT:.0f} s: {
stream\_url}",
21 )

23 with \_cameras\_lock:
24 current = \_cameras.get(stream\_url)
25 if current is None:
26 raise HTTPException(
27 status\_code=503,
28 detail="Camera was removed during startup"
29 )
30 count, connected = current.count, current.connected

32 return {
33 "camera": stream\_url,
34 "people\_count": count,
35 "connected": connected,
36 }


\textit{Listing 5.39: /current-count endpoint: full production implementation with post-cleanup safety}

check (droidip.py, lines 251--290).


The production endpoint differs from the prototype in three signiﬁcant ways. First,
it accepts an ip string parameter (and optional port integer) rather than a cam\_id inte-
ger, enabling it to serve any camera format supported by camera\_sources.py. Second,
the ready timeout is 15 seconds rather than 10, to accommodate RTSP cameras that take
longer to negotiate stream parameters. Third, it adds a post-cleanup safety check after
ready\_event.wait(): even after the event ﬁres, the cleanup worker could theoretically



delete the camera from the registry between the wait() returning and the second lock ac-
quisition. The \_cameras.get(stream\_url) call inside the second lock acquisition han-
dles this race --- if current is None, a 503 is returned rather than an AttributeError
\begin{lstlisting}
from accessing current.count on a NoneType object.
\end{lstlisting}

1 @app.get("/health")
2 def health():
3 """Return service health and basic stats."""
4 with \_cameras\_lock:
5 active = len(\_cameras)
6 return {
7 "status": "ok",
8 "active\_cameras": active,
9 "model\_loaded": \_model is not None,
10 }

12 @app.get("/")
13 def root():
14 return {"status": "running", "message": "People Counter API v2"}

16 if \_\_name\_\_ == "\_\_main\_\_":
17 uvicorn.run(app, host="0.0.0.0", port=8000)



\textit{Listing 5.40: /health and / endpoints, plus entry-point uvicorn launch (droidip.py, lines 292--322).}



/health is a standard operational endpoint that the .NET Core backend can poll to
conﬁrm the AI service is running. It reports three pieces of information: status ``ok'' (al-
ways, as long as the process is running); active\_cameras, the current count of registered
camera streams; and model\_loaded, a boolean indicating whether get\_model() has been
called at least once. The backend can use model\_loaded=false as an early-startup indi-
cator and delay its own camera polls until the model is loaded. The / root endpoint is a
minimal liveness check --- its only purpose is to return a 200 OK with a recognisable body,
conﬁrming the FastAPI application is running and accepting connections.

uvicorn.run(app, host="0.0.0.0", port=8000) binds the ASGI server to all
network interfaces on port 8000. host="0.0.0.0" is necessary for the .NET Core backend
to reach the service from a different process on the same machine or a different machine on
the local network; host="127.0.0.1" would restrict access to the same machine only. In



a production deployment, a reverse proxy (nginx, Caddy) should sit in front of uvicorn to
provide TLS termination and rate limiting.



\section{Administration and Conﬁguration Interface}


The Angular dashboard's administrative screens are where the hardware and AI services
described above are provisioned and where accounts are managed. This section presents all
three conﬁguration screens with annotated screenshots and traces each UI element back to
the relevant functional requirement and the code already discussed.




Figure 5.1: Devices page showing Esp32-Lab-D7 Active at 192.168.1.7 --- the same IP embedded
in the ﬁrmware's serverUrl constant.




\subsection{Add Lab --- Laboratory Provisioning Form}


The Add Lab form (Figure 5.2) is the administrative entry point for bringing a new laboratory into the GreenLab system. It collects exactly the information a new laboratory needs
and nothing more. Each ﬁeld is described below with its type, default value, constraint, and
connection to the backend and ﬁrmware.

This single form is the operational proof of NFR-SCAL-02's requirement that adding
a laboratory requires no schema or architecture change: every ﬁeld maps to an existing
database column or optional conﬁguration value, and creating a lab here does not require
touching the AI service, the ﬁrmware, or the database schema. It requires only ﬁlling in this
form and pairing one physical ESP32 with the resulting device key. The Create lab button





Figure 5.2: Add Lab: provisioning form for a new smart laboratory.


triggers a POST to the /api/labs endpoint with the form values serialised as JSON; the
Cancel button discards the form and navigates back to the Labs list.



\subsection{Devices --- Hardware Registry}





Figure 5.3: Devices: read-side registry showing Esp32-Lab-D7 Active at 192.168.1.7.

The Devices screen (Figure 5.3) is the read side of the provisioning step. Once a lab



Table 5.4: Add Lab form ﬁelds and their backend/ﬁrmware connections.

Field Type Default Purpose and code connection
Lab name Text, re- (none) Human-readable identiﬁer shown on all dash-
quired board views. Maps to the LabName column in
the Labs table.
Device key Text, op- (none) Credential used to pair the ESP32. The
(optional) tional ﬁrmware sends this in the token-exchange
POST to /api/devices/token alongside
labId and deviceId.
Light threshold Integer, 2000 Server-side copy of LIGHT\_THRESHOLD.
required Matches the ﬁrmware constant (Section 5.2.1).
Not yet fetched by the ﬁrmware at boot (NFR-
MAIN-01 gap).
Fan-on temper- Integer, 28 Server-side copy of the fan-on threshold.
ature ($^\circ$C) required Matches the ﬁrmware's implicit comparison
(temp >= 28 → fanOn). Same NFR-MAIN-
01 gap.
CV lab code Text, op- (none) Associates the lab with a computer-vision track-
(optional) tional ing context. Used by the AI service as a label
for camera registry keys.
Camera RTSP Text, op- (none) Passed to camera\_sources.py's
URL (optional) tional resolve\_source(). Accepts usb:N,
http/https/rtsp URL, or bare IPv4 (all ﬁve
formats from Section 5.3.2.2).




has been created and a device key issued, the corresponding ESP32 appears here with ﬁve
data columns: LAB (Lab D7), DEVICE (Esp32-Lab-D7), TYPE (ESP32 Controller), IP
(192.168.1.7), and STATUS (Active). The Active status is maintained by the ﬁrmware's
token lifecycle described in Section 5.2.4 --- a device that fails to authenticate or stops
sending events will have its status ﬂipped to Inactive by the backend after a conﬁgurable
idle timeout.

The IP address 192.168.1.7 matches the serverUrl, lightApi, and TokenGen constants embedded in the ﬁrmware source (Listing 5.3). This screen is therefore the backend's
own evidence that the ﬁrmware is successfully reaching the correct host --- if the IP were
wrong, the ﬁrmware's token exchange would fail, the device would never authenticate, and
the Status column would show Inactive rather than Active.








Figure 5.4: Users: account list with all four registered accounts across three roles.



\subsection{Users --- Account List}


The Users screen (Figure 5.4) shows all registered accounts in a tabular layout with ﬁve
columns: NAME, EMAIL, ROLE, STATUS, and ACTIONS (Edit / Delete). Four accounts are currently registered: Admin User (admin@greenlab.local, Admin, Active),
mohamed eldodge (mahammednasser0@gmail.com, Teaching Assistant, Active), TA User
(assistant@greenlab.local, Teaching Assistant, Active), and Teacher User (teacher@greenlab.local,
Teacher, Active). All four are Active, meaning their accounts have not been deactivated via
the Active account checkbox on the Add user form.

Three roles are in use: Admin, Teacher, and Teaching Assistant. These map to the
access-control tiers of the backend: Admin accounts can access all endpoints including
user management and lab provisioning; Teacher accounts can access monitoring and session management but not user administration; Teaching Assistant accounts are a reﬁnement
introduced during implementation (Section 5.4.4) that sits between Teacher and Admin in
terms of permissions.



\subsection{Add User --- Account Creation Form}


The Add user form (Figure 5.5) collects four ﬁelds: Full name, Email, Role (a dropdown
with options Teacher, Teaching Assistant, and Admin), and Password. The Active account
checkbox is checked by default, which activates the account immediately upon creation.





Figure 5.5: Add user: account creation form with role assignment and active-status toggle.


Unchecking it creates a provisioned-but-inactive account --- useful for pre-creating accounts for staff who have not yet started.

The password ﬁeld implements FR-AC-04's requirement that passwords are never
stored or transmitted in plaintext: the ﬁeld accepts a value that is sent to the backend and
immediately hashed server-side using a cryptographic hash function (bcrypt in the current
.NET Core implementation) before storage. The ﬁeld is never pre-ﬁlled when the Edit
button is used on an existing account --- editing an account shows all ﬁelds except the password, which must be explicitly re-entered to be changed. This prevents password exposure
via browser autoﬁll or network snifﬁng of a form pre-populated with a stored value.

The Save button posts to /api/users with the form values as JSON. The Cancel
button discards the form. After a successful save, the user list is refreshed to show the new
account, and the form collapses back to a hidden state --- the same UX pattern used for the
Add Lab form.




\section{Monitoring Dashboard Implementation}


This section covers the three screens that an instructor or administrator interacts with dayto-day: the login screen, the system-wide overview dashboard, and the per-laboratory monitoring view.






\subsection{Login Screen}





Figure 5.6: Login: the credential entry point implementing UC-01 and FR-AC-01.


The login screen (Figure 5.6) implements UC-01 and FR-AC-01 directly. It is a splitpanel layout: the left panel displays the GreenLab brand identity (``Powering a sustainable future''), a marketing tagline (``Manage your smart energy grid with precision and elegance''), and two summary metric cards (98% Efﬁciency and Opti Mode); the right panel
contains the authentication form. This layout follows a widely-used pattern for SaaS login
pages that reinforces brand values while presenting the credential entry form.

The credential form collects Email Address and Password, both shown with iconadorned input ﬁelds. The Password ﬁeld has a visibility toggle (the eye icon on the right)
that switches the input type between ``password'' (asterisks) and ``text'' (plaintext), helping
users who mistype their password without a show-password option. Forgot Password? is a
link to a password-reset ﬂow (not shown separately in this chapter). The Login → button
submits the credentials to the backend authentication endpoint. Alternatively, users can
continue with Google or Apple, which initiate OAuth2 ﬂows that ultimately produce the
same authenticated, role-bearing session token. The language toggle in the top-right corner
(English / Arabic) is an internationalisation affordance that allows Arabic-speaking staff to
use the system in their preferred language.

On successful authentication, the Angular router navigates to the Dashboard (Section 5.5.2). On failure (wrong password, inactive account, network error), an error message
is shown inline below the Login button without revealing whether the email or password



was speciﬁcally wrong --- a standard security practice to prevent user enumeration.



\subsection{System Dashboard}





Figure 5.7: Dashboard: weekly occupancy trend, active alerts panel, and recent activity feed.


The administrator's landing dashboard (Figure 5.7) is the primary daily-use screen for
lab managers. It implements FR-MD-03 (system-wide status at a glance) and FR-FH-03
(faults visible without navigation). The layout is divided into three zones.

The upper zone occupies most of the screen width and shows a weekly occupancy
trend chart. The x-axis spans the seven days of the current week (Mon through Sun), and
the y-axis represents the aggregate occupancy index --- a composite derived from per-lab
occupancy counts weighted by session status. The curve peaks on Thursday and Friday,
consistent with a typical university schedule where more labs are in use mid-to-late week.
Hollow circles on the curve indicate speciﬁc measurement points, and the ﬁlled circle on
Thursday marks the current day's peak. The right side of the upper zone shows a bar chart
comparing the relative occupancy index (0--24 range) across three lab categories: Network
Lab, Software Lab, and Hardware Lab.

The lower-left zone is the Active alerts panel, currently showing two alerts: ``Occupancy outside session --- Software Lab B: Last crowd count: 15. Check schedule or
overrides.'' (3 hours ago) and ``Occupancy outside session --- Hardware Lab C: Last crowd
count: 8.'' (4 hours ago). Each alert has a dismiss button (×) in the top-right corner. These



alerts are the implementation of the ``Occupancy outside session'' fault condition identiﬁed
in Section 5.4.4 --- a condition where the AI service reports non-zero occupancy in a lab
that the session schedule shows as unbooked. This alert type does not appear in the Chapter 3 FR list by name and is recommended for formal inclusion as FR-FH-04 in a future
revision.

The lower-right zone is the Recent activity feed, showing timestamped per-lab events:
``Network Lab A: occupancy 24 · In session'' (3 hr ago), ``Software Lab B: occupancy 15
· Out of session'' (3 hr ago), ``Hardware Lab C: occupancy 8 · Out of session'' (4 hr ago).
A green dot indicates ``In session''; a grey dot indicates ``Out of session''. The Tip card at
the bottom suggests a speciﬁc administrative action: ``Open Network Lab A to see devices
and thresholds from the API when signed in as admin.'' This tip is context-aware --- it is
triggered by the presence of a lab with the highest occupancy and provides a direct action
link, not generic advice.



\subsection{All Labs --- Per-Laboratory View}





Figure 5.8: All Labs: Lab D7 shown Idle with 4/30 kWh energy usage and three device-state pills.


The All Labs screen (Figure 5.8) is the per-laboratory monitoring view implementing
FR-MD-01. It shows one status card per registered laboratory. The Lab D7 card displays:
a status pill (Idle, shown in amber) in the top-right corner; an energy usage bar labelled
``ENERGY USAGE'' with a green progress bar at approximately 13% ﬁll, captioned ``4 /
30 kWh''; three device-state pills in the lower-left corner, all reading ``IDLE'' in amber; a



last-updated timestamp (``36 d ago'') in the lower-right corner; and an ``Add Labs +'' button
in the top-right corner of the page for quick access to the Add Lab form.

The three device-state pills correspond to the three relay channels visible in the dashboard's data model --- light, fan, and a third channel. As noted in Section 5.2.4, only
two relay channels (light and fan) are physically wired in the current ﬁrmware build, but
the dashboard already renders three slots. The third slot reading IDLE is consistent with
relay3On never being set to true in the backend's response, since the ﬁrmware logs it but
does not drive a GPIO pin for it.

Below the lab card, an AI Recommendation card reads: ``Review session schedules
and occupancy in the admin API to tune savings.'' This recommendation is generated by
the backend's analytics engine, which correlates session schedules with occupancy data to
identify conﬁguration changes that would reduce unnecessary energy consumption. It is
an emergent feature surfaced from occupancy data rather than from a separately speciﬁed
requirement, and is noted here as a candidate for formal speciﬁcation in a future project
iteration.



\section{Implementation Traceability Matrix}


The table below maps every functional and non-functional requirement identiﬁer from
Chapter 3 to the speciﬁc section, listing, ﬁgure, or gap note in this chapter where it is
evidenced.

Table 5.5: Implementation traceability matrix.


Req. ID Requirement Summary Evidence in Ch. 5 Status

FR-AC-01 Device/user authentica- §5.5.1 Fig. 5.6; §5.2.4 Implemented
tion on login Listings 5.8--5.10
FR-AC-02 All non-login endpoints §5.2.4 (device); §5.5.1 Implemented
require valid JWT (users); §5.4.3
FR-AC-03 Role-checked JWT on ev- §5.4.3--5.4.4 three-role Implemented
ery endpoint model







Req. ID Requirement Summary Evidence in Ch. 5 Status

FR-AC-04 Passwords hashed server- §5.4.4 Add user form Implemented
side; never redisplayed analysis
FR-EE-02 Debounce / minimum- §5.2.6 Listing 5.14 (edge- Implemented
interval between motion trigger + COOLDOWN)
reports
FR-FH-01 Timeout-based fault de- §5.3.3.9 Listing 5.39 Implemented
tection at AI boundary READY\_TIMEOUT 503
FR-FH-02 Hold last relay state dur- §5.2.6 Listing 5.15 fall- Partial
ing connectivity fault back (re-evaluates, not
freezes)
FR-FH-03 Faults visible on dash- §5.5.2 Fig. 5.7 Active Implemented
board without navigation alerts panel
FR-MD-01 Live sensor-state and §5.5.3 Fig. 5.8 device- Implemented
relay-state view per lab state pills
FR-MD-03 System-wide status at a §5.5.2 Fig. 5.7 full dash- Implemented
glance board layout
FR-OD-01 Person detection $\ge$0.5 §5.3.3.7 Listing 5.36; Implemented
conﬁdence, $\le$5 s cadence FRAME\_SKIP=3
FR-OD-02 Empty/Low/Moderate/High Dashboard occupancy la- Implemented
occupancy classiﬁcation bels Figs. 5.7--5.8
FR-SI-04 Device re-authenticates §5.2.5 Listing 5.12 retry Implemented
after 401 block
NFR- Firmware thresholds Hardcoded constants Not yet imMAIN-01 fetched from API at §5.2.1 --- not fetched plemented
runtime
NFR-PERF- AI service sampling §5.3.3.6 Implemented
03 within 5-second cadence FRAME\_SKIP=3 at
30 fps → 10 inferences/s







Req. ID Requirement Summary Evidence in Ch. 5 Status

NFR-SCAL- Adding a lab requires §5.4.1 Fig. 5.2 Add Lab Implemented
02 no schema/architecture form
change




\section{Worked End-to-End Example: One Motion Event Across}

All Four Tiers

This section traces a single real-world scenario --- a student entering Lab D7 at 09:15 on
a Thursday morning while a session is scheduled --- through all four system tiers. The
JSON payloads shown are representative of those exchanged by the running system on the
development network.




\subsection{Tier 1 --- ESP32 Sensor Reads (loop() iteration N)}


The main loop iteration begins. readDistanceCM() ﬁres the HC-SR04 and measures
a round-trip of 2,235 microseconds, corresponding to a distance of 2235 × 0.034/2 =
38 cm. Since 38 > 0 and 38 $\le$ 50 (DISTANCE\_THRESHOLD), isNear evaluates to true.
The previous iteration had isNear = false (the doorway was unoccupied), so wasNear
= false. The condition isNear \&\& !wasNear \&\& (millis() - lastToggleTime >
\begin{itemize}
  \item is satisﬁed --- this is a rising edge with sufﬁcient cooldown elapsed.
\end{itemize}

readLight() returns 1850, indicating a moderately dim room. dht.readTemperature()
\begin{lstlisting}
returns 26.4 and dht.readHumidity() returns 61.0; neither is NaN so the isnan() guard
\end{lstlisting}

passes. ensureTokenValid() checks the cached token and ﬁnds millis() = 1,200,000 ms
(20 minutes since boot) is less than tokenExpireMillis = 3,600,000 ms (60 minutes
since boot) --- token is valid, no network call. WiFi.status() returns WL\_CONNECTED.
sendMotionRequest() is called.







\subsection{Tier 1 → Tier 2: HTTP POST to the ASP.NET Core Backend}


sendMotionRequest() constructs the following JSON body by manual string concatena-
tion:

1 POST http://192.168.1.7:5076/api/devices/1/event HTTP/1.1
2 Authorization: Bearer eyJhbGciOiJIUzI1NiIsInR5cCI6IkpXVCJ9.[payload].[signature]
3 Content-Type: application/json

5 {
6 "motionDetected": true,
7 "distanceCm": 38,
8 "ldrValue": 1850,
9 "temperatureCelsius": 26.4,
10 "humidity": 61.0
11 }



\textit{Listing 5.41: Full HTTP POST sent by the ESP32 on a rising-edge motion event.}



The backend receives this request, validates the JWT signature and expiry, identi-
ﬁes the calling device as Esp32-Lab-D7 associated with Lab D7, and begins evaluating
the automation rules. The session schedule shows a session running from 08:00 to 12:00
today, so session-based rules are active. The backend polls the AI service for the cur-
rent occupancy count (Tier 3, below) and receives people\_count = 1. The backend
evaluates: ldrValue (1850) < lightThreshold (2000) → room is dark → lightOn
= true; temperatureCelsius (26.4) < fanOnTemp (28) → room is not hot → fanOn
= false. motionIgnored = false because the event falls within a scheduled session.
occupancyCount = 1.



\subsection{Tier 2 → Tier 3: AI Service Poll (concurrent with backend evalu-}

ation)

The backend sends a GET request to the Python AI service:

1 GET http://127.0.0.1:8000/current-count?ip=192.168.1.3 HTTP/1.1



\textit{Listing 5.42: Backend polling the AI service for current occupancy.}





The AI service's camera loop thread for http://192.168.1.3:4747/video has
been running continuously. On the most recent processed frame (frame index 72, the 24th
inference run since the camera started), ByteTrack assigned track\_id = 7 to the enter-
ing student. The tracked\_ids set contains {7}, so \_run\_inference() returned 1. The
CameraState for this camera has count = 1 and connected = True. The AI service
responds:
1 HTTP/1.1 200 OK
2 Content-Type: application/json


4 {
5 "camera": "http://192.168.1.3:4747/video",
6 "people\_count": 1,
7 "connected": true
8 }


\textit{Listing 5.43: AI service response conﬁrming one tracked person in frame.}





\subsection{Tier 2 → Tier 1: Backend Response}


The backend constructs its response and returns it to the ESP32:
1 HTTP/1.1 200 OK
2 Content-Type: application/json

4 {
5 "lightOn": true,
6 "fanOn": false,
7 "relay3On": false,
8 "relay4On": false,
9 "motionIgnored": false,
10 "occupancyCount": 1
11 }


\textit{Listing 5.44: Backend JSON response driving ESP32 relay actuation.}



sendMotionRequest() receives HTTP 200 with this body. The code is not 401,
so no retry is needed. getString() returns the JSON body, which is returned from
sendMotionRequest() to the motion-event block in loop().




\subsection{Tier 1 --- Relay Actuation}


Back in loop(), response is the JSON body above. deserializeJson() succeeds.
occupancy = 1, so the room-empty override is skipped. lightOn = true, lastLightState
= false → lightOn != lastLightState is true → digitalWrite(RELAY\_PIN, LOW)
closes the relay → light turns ON. lastLightState = true. fanOn = false, lastFanState
= false → no relay write needed. relay3On = false, relay4On = false → logged to
serial only. motionIgnored = false → no early return. lastToggleTime = millis().
wasNear = true (set at the bottom of loop()). The entire round-trip from the HC-SR04
ping to the light switching on takes under 600 ms on the local network.



\subsection{Tier 4 --- Dashboard Update}


The Angular dashboard's polling interval (conﬁgured at 5 seconds) ﬁres a GET to /api/labs/d7/status.
The backend returns the updated lab state including occupancy = 1, lightOn = true,
fanOn = false, sessionStatus = "In session". The Recent activity feed updates to
show ``Lab D7: occupancy 1 · In session'' with a green dot. No alert is raised because the
occupancy occurred within a scheduled session.



\section{Known Limitations and Technical Debt}


Throughout this chapter, implementation gaps and deviations from Chapter 3 speciﬁcations
have been noted as they arose. The table below consolidates all seven items into a single
reference, including the affected requirement, the root cause, and a concrete recommended
ﬁx.






Table 5.6: Known limitations and technical debt.


Item Description Req. Recommended ﬁx

Firmware LIGHT\_THRESHOLD (2000) and NFR- Add a conﬁg end-
thresholds fan-on temperature (28$^\circ$C) MAIN- point; call it at boot;
hardcoded are compile-time constants. 01 cache in EEPROM
Changing a threshold requires for ofﬂine use.
reﬂashing the device. No
/api/devices/{id}/config
endpoint exists yet.
Two relay relay3On and relay4On are ArchitectureAdd \#define
channels parsed and logged but not con- diagram RELAY\_PIN3/4; add
unwired nected to GPIO outputs. Only actuation logic ---
RELAY\_PIN and RELAY\_PIN2 no backend change
are physically wired. Hard- needed.
ware not yet expanded beyond
two channels.
Local fallback Fallback logic re-evaluates FR-FH- Store last backend-
re-evaluates LDR rather than holding the 02 commanded light
vs. freezes last backend-commanded state; restore it on
state as FR-FH-02 speciﬁes. fallback rather than
Designed for simplicity over re-evaluating.
strict spec compliance.
Teaching The role was added during FR-AC- Add Teaching Assis-
Assistant role implementation but is absent 02/03 tant as a named actor
unmodelled from the Chapter 3 use case in the use case dia-
model. Discovered during im- gram and stakeholder
plementation as a real organi- analysis.
sational need.







Item Description Req. Recommended ﬁx

``Occupancy The dashboard raises this alert FR-FH Write back as FR-
outside ses- when occupancy is detected (future) FH-04 in a revised
sion'' alert outside scheduled hours. No Chapter 3.
unspeciﬁed FR ID exists for it. Emergent
feature identiﬁed during dash-
board implementation.
Device iden- labId and deviceId are in- NFR- Read
tity compiled teger literals in the ﬁrmware. MAIN- labId/deviceId
in Deploying to a second lab re- 01 from EEPROM,
quires recompiling. No provi- provisioned via
sioning API was designed for QR code or BLE
device self-registration. commissioning.
FRAME\_SKIP At variable camera frame NFR- Replace
is count- rates, the effective inference PERF-03 FRAME\_SKIP with a
based not cadence varies. At 15 fps minimum-elapsed-
time-based FRAME\_SKIP=3 gives 5 infer- time gate based on
ences/s; at 5 fps it gives only time.monotonic().
1.7/s.




\section{Chapter Summary}


This chapter presented the as-built GreenLab system in full implementation detail across
its two directly-implemented tiers --- the ESP32 hardware layer and the Python AI processing service --- and through annotated screenshots for the Angular administration and
monitoring interface.

Section 5.2 covered the ESP32 ﬁrmware in six subsections. Section 5.2.1 documented
all library includes, pin assignments, and global constants, explaining the rationale behind each GPIO choice and the consistency gap between hardcoded thresholds and the
Add Lab form defaults. Section 5.2.2 presented the three sensor-reading functions with
physics-level explanations of the HC-SR04 time-of-ﬂight calculation, the ESP32 ADC



non-linearity caveat, and the DHT11 NaN guard. Section 5.2.3 documented the four local
HTTP server handlers that enable direct relay control from the dashboard without a backend
round-trip, and explained the active-low relay convention and the importance of updating
the state trackers inside each handler. Section 5.2.4 presented the complete JWT token
lifecycle --- handleTokenResponse(), ensureTokenValid(), and refreshToken() ---
with explanations of the 50-minute assumed expiry, the millis() overﬂow edge case,
and the security note on plaintext credentials in the local network context. Section 5.2.5
presented sendMotionRequest() with line-by-line explanation of the pre-ﬂight guards,
manual JSON construction, 401 retry logic, and the http.end() necessity. Section 5.2.6
presented setup() and loop() in full, covering safe-state relay initialisation, non-blocking
Wi-Fi watchdog, edge-triggered motion detection, local fallback with differential update,
JSON response parsing, room-empty override, and differential relay actuation.

Section 5.3 covered the Python AI service in three ﬁles. Section 5.3.1 presented
cout\_code.txt --- the single-camera prototype --- covering the double-checked locking
singleton model loader, the two-attempt VideoCapture open strategy, the person-counting
generator expression, the ready-event signalling pattern, and the lock-release-before-wait
pattern in the endpoint. Section 5.3.2 presented camera\_sources.py in full --- the ﬁvebranch resolve\_source() dispatch, the frozenset for O(1) extension lookup, the IPv4
regex, create\_capture()'s intentional no-exception design, source\_label()'s idempotent key generation, and the two private parsing helpers with their speciﬁc validation strategies. Section 5.3.3 presented droidip.py --- the production service --- covering structured
logging, the CameraState dataclass and monotonic last\_access, \_update\_state()'s
cleanup-race guard, the dual deregistration checks in \_camera\_loop(), FRAME\_SKIP framecount sampling, the exception guard around inference, ByteTrack's persist=True and
classes=[0] arguments, \_ensure\_camera\_started()'s atomic registration, \_cleanup\_worker()'s
separate-list-before-delete pattern, and the /current-count post-cleanup safety check.

Sections 5.4 and 5.5 covered the Angular dashboard through annotated screenshots:
the Add Lab form with a per-ﬁeld analysis table, the Devices registry and its connection
to the ﬁrmware's IP constants, the Users list showing the three-role model, the Add user
form and its FR-AC-04 password handling, the Login screen with its OAuth2 alternatives,
the system Dashboard with its three information zones and the ``Occupancy outside session'' emergent alert, and the All Labs per-laboratory view with its device-state pills and AI



Recommendation card.

Section 5.6 provided a complete requirement traceability matrix. Section 5.7 traced
one complete motion event through all four tiers using representative JSON payloads. Section 5.8 consolidated seven implementation gaps with root causes and concrete recommended ﬁxes.

Read together with Chapter 3, this chapter closes the loop from requirement to running
system. Every functional and non-functional requirement area introduced in Chapter 3
now has at least one code listing, ﬁgure, or honest gap note in this chapter. The seven
gaps identiﬁed --- hardcoded thresholds, two unwired relay channels, fallback behaviour
deviation, unmodelled role, unspeciﬁed alert type, compiled-in device identity, and framecount-based rather than time-based sampling --- are recorded precisely here so that they
can be addressed deliberately in the testing and conclusion chapters rather than discovered
unexpectedly during examination.




