\chapter*{Abstract}

GreenLab is an intelligent energy management system designed for computer science faculty laboratories. The system combines embedded sensing, automated appliance control, computer-vision occupancy estimation, backend decision logic, role-based monitoring, and manual override through a web dashboard. 

The hardware layer is centred on an ESP32 microcontroller connected to environmental sensors (LDR for natural-light sensing, DHT11 for temperature and humidity, HC-SR04 ultrasonic for entry and activity detection) and a 4-channel relay module for controlling lights and fans. The artificial-intelligence layer uses camera frames and a YOLO-based model to estimate the number of occupants in the laboratory. The backend layer receives sensor and occupancy data, applies operational rules, stores audit records, and exposes secure REST endpoints. The frontend layer presents real-time status, laboratory assignments, manual controls, and administrative actions through an Angular single-page application dashboard.

This project addresses the gap between purely manual control and purely schedule-based automation by providing an integrated cyber-physical system that connects embedded sensing, occupancy detection, rule-based automation, supervised override, and administrative governance in one coherent platform. The system is designed using a four-tier architecture that separates hardware sensing, AI processing, backend decision-making, and dashboard presentation, enabling each tier to evolve independently while maintaining well-defined interface contracts.
