% ─────────────────────────────────────────────────────────────
% References
% ─────────────────────────────────────────────────────────────
\chapter*{References}
\label{ch:references}
\addcontentsline{toc}{chapter}{References}

\begin{thebibliography}{99}

\bibitem{challa2025}
R.~Challa et~al., ``Multimodal classroom occupancy detection for HVAC optimization,'' in \textit{Proc. IEEE Int. Conf. Smart Buildings}, 2025.

\bibitem{chaudhari2024}
A.~Chaudhari, Y.~Xiao, J.~Cheng, and G.~Li, ``A comprehensive survey of occupancy detection technologies for smart buildings using IoT sensors,'' \textit{IEEE Internet Things J.}, vol.~11, no.~5, pp.~8234--8256, 2024.

\bibitem{zhang2026}
L.~Zhang et~al., ``Indoor head detection with improved lightweight YOLOv8 on SCUT-HEAD dataset,'' \textit{Appl. Sci.}, vol.~16, no.~1, p.~335, 2026.

\bibitem{waghchoure2025}
S.~Waghchoure et~al., ``IoT-based classroom energy-saving system using PIR motion sensing and camera-based detection,'' in \textit{Proc. Int. Conf. IoT Smart Environ.}, 2025.

\bibitem{wibowo2023}
A.~Wibowo et~al., ``IoT-based energy management system for university laboratory environments,'' \textit{J. Sustain. Comput.}, vol.~14, pp.~100--115, 2023.

\bibitem{ahmed2020}
M.~Ahmed et~al., ``IoT-based classroom automation system with sensor-driven control and web dashboard,'' in \textit{Proc. Int. Conf. Smart Technol.}, 2020.

\bibitem{ingole2024}
R.~P.~Ingole, A.~K.~Meshram, S.~J.~More, N.~S.~Dhumne, and S.~M.~Gawarle, ``Automation of IoT-based classroom: Integrating light, fan control, temperature monitoring, and entry management using ESP32 microcontroller,'' \textit{IRJMETS}, vol.~6, no.~5, pp.~4642--4645, 2024.

\bibitem{smarthaus}
SmartHaus Project, ``SmartHaus: Open-source home automation system with Angular dashboard and REST API.'' [Online]. Available: \url{https://github.com/smarthaus/angular-iot-dashboard}

\bibitem{song2024}
Y.~Song et~al., ``Impact of occupancy behaviour on building energy efficiency,'' 2024.

\bibitem{park2018}
J.~Y.~Park, T.~Dougherty, and Z.~Nagy, ``A bluetooth based occupancy detection for buildings,'' in \textit{Proc. Building Perform. Anal. Conf. SimBuild (IBPSA)}, 2018.

\bibitem{yang2018}
J.~Yang, A.~Pantazaras, K.~A.~Chaturvedi, and A.~K.~Chandran, ``Comparison of different occupancy counting methods for single system-single zone applications,'' \textit{Energy Build.}, vol.~172, pp.~221--234, 2018.

\bibitem{chen2018}
Z.~Chen, Y.~Wang, and H.~Liu, ``Unobtrusive sensor-based occupancy facing direction detection and tracking using advanced machine learning algorithms,'' \textit{IEEE Sensors J.}, vol.~18, pp.~6360--6368, 2018.

\bibitem{longo2019}
E.~Longo, A.~E.~C.~Redondi, and M.~Cesana, ``Accurate occupancy estimation with WiFi and bluetooth/BLE packet capture,'' \textit{Comput. Netw.}, vol.~163, p.~106876, 2019.

\bibitem{hobson2019}
B.~W.~Hobson et~al., ``Opportunistic occupancy-count estimation using sensor fusion: a case study,'' \textit{Build. Environ.}, vol.~159, p.~106154, 2019.

\bibitem{mo2019}
J.~Mo, ``A comprehensive framework on occupancy indicators for optimizing building operation,'' \textit{Energy Build.}, vol.~182, pp.~110--123, 2019.

\bibitem{li2019}
X.~Li, ``An IoT sensor setup for tracking occupancy patterns and profiles in office spaces,'' \textit{Build. Environ.}, vol.~154, pp.~22--35, 2019.

\bibitem{tekler2020}
Z.~D.~Tekler et~al., ``A scalable bluetooth low energy approach to identify occupancy patterns and profiles in office spaces,'' \textit{Build. Environ.}, vol.~171, p.~106681, 2020.

\bibitem{park2020}
S.~Park, ``Demand-driven load prediction models using continuous real-time occupancy estimation,'' \textit{Energy Build.}, vol.~204, p.~109482, 2020.

\bibitem{alishahi2021}
N.~Alishahi, M.~Nik-Bakht, and M.~M.~Ouf, ``A framework to identify key occupancy indicators for optimizing building operation using WiFi connection count data,'' \textit{Build. Environ.}, vol.~200, p.~107936, 2021.

\bibitem{mukhopadhyay2021}
B.~Mukhopadhyay, ``Indoor localization using analog output of pyroelectric infrared sensors,'' \textit{IEEE Sensors J.}, vol.~21, no.~15, pp.~17053--17063, 2021.

\bibitem{mitra2022}
D.~Mitra, D.~Malekpour~Koupaei, and K.~Cetin, ``Occupancy data sensing, collection, and modeling for residential buildings,'' in \textit{Renewable Energy for Buildings}. Springer, 2022, pp.~103--121.

\bibitem{kim2023}
M.-L.~Kim, K.-J.~Park, and S.-Y.~Son, ``Occupancy-based energy consumption estimation improvement through deep learning,'' \textit{Sensors}, vol.~23, p.~2127, 2023.

\bibitem{khan2024}
I.~Khan, ``Occupancy prediction in IoT-enabled smart buildings: technologies, methods, and future directions,'' \textit{Sensors}, vol.~24, no.~11, p.~3276, 2024.

\bibitem{alturjman2024}
F.~Al-Turjman, ``Data transmission optimizations in occupancy monitoring and smart HVAC networks,'' \textit{IEEE Internet Things J.}, vol.~11, no.~2, pp.~1432--1445, 2024.

\bibitem{tanaka2024}
H.~Tanaka and Y.~Mori, ``Advanced occupancy sensing configurations via low-power ambient network nodes,'' \textit{J. Build. Eng.}, vol.~84, p.~108112, 2024.

\bibitem{garcia2024}
A.~R.~Garcia, ``Multi-zone occupancy classification models utilizing distributed environmental sensor networks,'' \textit{Energy Build.}, vol.~310, p.~114095, 2024.

\bibitem{green2024}
M.~S.~Green and L.~J.~Taylor, ``Commercial office spatial profiling through integrated BLE and localized CO$_2$ metrics,'' \textit{Build. Environ.}, vol.~250, p.~111190, 2024.

\bibitem{zhao2024}
K.~Zhao and T.~Fletcher, ``Deep learning applications for predictive thermal comfort and occupant positioning,'' \textit{Appl. Energy}, vol.~358, p.~122602, 2024.

\bibitem{patel2024}
S.~G.~Patel, ``Real-time non-intrusive spatial occupancy counting protocols based on device-free passive sensing,'' \textit{IEEE Trans. Smart Grid}, vol.~15, no.~3, pp.~2451--2463, 2024.

\bibitem{lindqvist2025}
J.~Lindqvist and O.~Nilsson, ``Thermal camera array fusion architectures for high-density public lounge occupancy profiling,'' \textit{Sensors}, vol.~25, no.~4, p.~1189, 2025.

\bibitem{gomez2025}
R.~V.~Gomez, ``A comparative survey on edge-computed ambient intelligence for dynamic building automation systems,'' \textit{Autom. Constr.}, vol.~169, p.~105820, 2025.

\bibitem{nakajima2025}
Y.~Nakajima and S.~Sato, ``Decentralized room occupant counting via distributed acoustic signature deep neural networks,'' \textit{Appl. Sci.}, vol.~15, no.~2, p.~804, 2025.

\bibitem{dupont2025}
C.~M.~Dupont, ``Evaluating privacy-preserving localized radar technology for spatial analytics in commercial real estate,'' \textit{Energy Build.}, vol.~332, p.~115201, 2025.

\bibitem{nguyen2025}
T.~H.~Nguyen and V.~P.~Tran, ``Graph neural network frameworks for multi-zone institutional building occupancy forecasting,'' \textit{Int. J. Energy Res.}, vol.~2025, pp.~1--18, 2025.

\bibitem{mutiset2022}
Mutiset et~al., ``Multi-stream deep neural network for identifying human activities using YOLOv3 network,'' \textit{Appl. Intell.}, vol.~52, pp.~147--152, 2022.

\bibitem{cervantes2020}
J.~Cervantes, F.~Garcia-Lamont, L.~Rodr\'{\i}guez-Mazahua, and A.~Lopez, ``A comprehensive survey on support vector machine classification: applications, challenges and trends,'' \textit{Neurocomputing}, vol.~408, pp.~18--35, 2020.

\bibitem{murray2021}
D.~G.~Murray, J.~Simsa, A.~Klimovic, and I.~T.~Indyk, ``tf.data: A machine learning data processing framework,'' 2021, arXiv:2101.12127.

\bibitem{zhou2021}
T.~Zhou, ``Vision-based tracking framework vs YOLO-only baselines in commercial buildings,'' \textit{J. Build. Eng.}, vol.~45, p.~103421, 2021.

\bibitem{wang2020}
L.~Wang, ``Machine learning architectures for multi-occupant environments,'' \textit{IEEE Trans. Smart Grid}, vol.~11, no.~4, pp.~3122--3134, 2020.

\bibitem{choi2023}
Choi et~al., ``Effectiveness of YOLOv5 algorithm within office environments and user approval tracking,'' \textit{Build. Environ.}, vol.~231, p.~110012, 2023.

\bibitem{meng2023}
Y.~Meng et~al., ``A CNN-based, end-to-end system for predicting dynamic occupancy loads in building spaces enabling real-time estimation,'' \textit{Build. Environ.}, vol.~228, p.~109815, 2023.

\bibitem{chen2023}
Y.~Chen, D.~Zhu, N.~Li, Y.~Zhou, and Y.~Bai, ``GET: Group equivariant transformer for person detection of overhead fisheye images,'' \textit{Appl. Intell.}, vol.~53, pp.~24551--24565, 2023.

\bibitem{meng2023b}
Y.~Meng, T.~Li, G.~Liu, S.~Xu, and T.~Ji, ``Real-time dynamic estimation of occupancy load and an air-conditioning predictive control method based on image information fusion,'' \textit{Build. Environ.}, vol.~235, p.~110212, 2023.

\bibitem{wu2024}
X.~Wu and M.~Jin, ``Lightweight cross-stage partial network architectures for edge-deployed human counting,'' \textit{Pattern Recognit. Lett.}, vol.~178, pp.~45--52, 2024.

\bibitem{kumar2024}
S.~Kumar and R.~Gupta, ``Robust multi-person tracking under severe occlusion using DeepSORT and transformer embeddings,'' \textit{Comput. Vis. Image Underst.}, vol.~239, p.~103901, 2024.

\bibitem{zhang2024}
L.~Zhang, ``An attention-guided feature pyramid network for multi-scale pedestrian identification in complex indoor scenery,'' \textit{IEEE Access}, vol.~12, pp.~18921--18934, 2024.

\bibitem{lee2024}
K.~H.~Lee, ``Real-time overhead person counting utilizing modified anchor-free structural detectors,'' \textit{J. Real-Time Image Process.}, vol.~21, no.~2, p.~43, 2024.

\bibitem{mitchell2024}
J.~P.~Mitchell, ``Optimizing computational layers in YOLO configurations for resource-constrained edge systems,'' \textit{Mach. Learn. Appl.}, vol.~15, p.~100518, 2024.

\bibitem{rossi2024}
M.~Rossi and G.~Bianchi, ``Privacy-by-design low-resolution vision networks for institutional crowd management,'' \textit{Int. J. Comput. Vis.}, vol.~132, no.~5, pp.~1102--1119, 2024.

\bibitem{park2024}
J.~H.~Park and S.~W.~Kim, ``Feature map scaling layers inside deep learning blocks for rapid low-altitude target processing,'' \textit{Neurocomputing}, vol.~580, p.~127410, 2024.

\bibitem{adebayo2024}
O.~S.~Adebayo, ``Automated dynamic bounding box refinement protocols using transformer-based visual tokens,'' \textit{IEEE Trans. Multimedia}, vol.~26, pp.~3105--3118, 2024.

\bibitem{rodriguez2024}
E.~F.~Rodriguez, ``Self-supervised feature extraction schemes via spatial transformer arrays for panoramic fisheye rectification,'' \textit{Image Vis. Comput.}, vol.~145, p.~104992, 2024.

\bibitem{elsayed2026}
A.~M.~El-Sayed, ``Optimized deep multi-frame target tracking metrics under dynamic lighting shifts,'' \textit{Vis. Comput.}, vol.~42, no.~1, pp.~215--231, 2026.

\bibitem{kodali2017}
R.~K.~Kodali and K.~S.~Mahesh, ``Low cost implementation of smart home automation,'' in \textit{Proc. IEEE ICACCI}, 2017, pp.~461--466.

\bibitem{marhoon2018}
H.~M.~Marhoon, M.~I.~Mahdi, E.~D.~Hussein, and A.~R.~Ibrahim, ``Designing and implementing applications of smart home appliances,'' \textit{Mod. Appl. Sci.}, vol.~12, no.~12, pp.~8--17, 2018.

\bibitem{dhanashri2020}
D.~V.~Dhanashri, D.~R.~Kamble, G.~V.~Bhosale, P.~D.~Bhosale, and P.~P.~Kale, ``Air pollution detection using IOT,'' \textit{J. Eng. Comput. Archit.}, vol.~10, no.~4, pp.~442--447, 2020.

\bibitem{simanungkalit2020}
Simanungkalit et~al., ``Temperature control of computer laboratory air conditioners using DHT11 and Arduino,'' \textit{J. Phys.: Conf. Ser.}, vol.~1424, p.~012054, 2020.

\bibitem{basheer2021}
Basheer et~al., ``Multi-sensor laboratory environment tracking system using motion, temperature, and luminosity sensors,'' \textit{Int. J. Comput. Appl.}, vol.~182, no.~12, pp.~33--38, 2021.

\bibitem{sowndarya2021}
P.~Sowndarya and N.~Saranya, ``Sensor based industrial kitchen foodstuffs monitoring system,'' \textit{Int. J. Comput. Commun. Inform.}, vol.~3, no.~1, pp.~26--43, 2021.

\bibitem{rao2021}
G.~S.~Rao, M.~B.~Shanmugi, D.~Pradeepa, and V.~U.~Kamali, ``Smart home automation system by integrating multiple sensors,'' \textit{IOP Conf. Ser.: Mater. Sci. Eng.}, vol.~1208, p.~012011, 2021.

\bibitem{anderson2024}
T.~N.~Anderson, ``ESP32 microarchitectures for localized edge processing and multi-peripheral sensor scheduling,'' \textit{IoT Interface Lett.}, vol.~8, pp.~112--119, 2024.

\bibitem{foster2024}
M.~J.~Foster, ``Automated load balancing algorithms for distributed campus power management via MQTT networks,'' \textit{IEEE Trans. Consum. Electron.}, vol.~70, no.~2, pp.~501--512, 2024.

\bibitem{silva2024}
L.~C.~Silva, ``Implementing localized firmware filters for mitigating signal bouncing in optocoupler-based entrance trackers,'' \textit{Microprocess. Microsyst.}, vol.~104, p.~104990, 2024.

\bibitem{jenkins2024}
H.~B.~Jenkins, ``Performance limits of low-cost infrared break-beam systems under inconsistent daylight interference,'' \textit{Opt. Lasers Eng.}, vol.~175, p.~107981, 2024.

\bibitem{kim2024}
Y.~D.~Kim and J.~Lee, ``Designing real-time smart schedule adaptors for university HVAC operations using real-time node polling,'' \textit{Energy Build.}, vol.~318, p.~114402, 2024.

\bibitem{prasad2024}
S.~Prasad, ``Low-power environmental instrumentation protocols inside academic institutional infrastructures,'' \textit{Sustain. Comput.: Inform. Syst.}, vol.~42, p.~100961, 2024.

\bibitem{patel2024b}
A.~M.~Patel, ``A cloud-managed multi-sensor hub for localized classroom lighting and student profile assessment,'' \textit{J. Netw. Comput. Appl.}, vol.~226, p.~103855, 2024.

\bibitem{mwangi2024}
F.~K.~Mwangi, ``Low-latency edge relay network designs for sudden occupant surges inside public institutions,'' \textit{Comput. Commun.}, vol.~219, pp.~88--99, 2024.

\bibitem{thompson2024}
G.~Thompson, ``Evaluating long-term calibration retention of low-cost DHT22 and DHT11 modules under variable student density,'' \textit{Measurement}, vol.~230, p.~114510, 2024.

\bibitem{davies2024}
J.~A.~Davies, ``Scalable multi-classroom control arrays utilizing integrated I2C and SPI peripheral configurations,'' \textit{IEEE Trans. Ind. Inform.}, vol.~20, no.~4, pp.~5890--5901, 2024.

\bibitem{fischer2025}
R.~M.~Fischer, ``A decentralized master-slave architecture for real-time room power control via low-power mesh nodes,'' \textit{Sens. Actuators A}, vol.~368, p.~115090, 2025.

\bibitem{nielsen2025}
P.~H.~Nielsen, ``Dynamic window actuator orchestrations linked with adaptive multi-sensor CO$_2$ triggers in smart schools,'' \textit{Build. Environ.}, vol.~272, p.~112501, 2025.

\end{thebibliography}
